---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: console
---

Separating biological signal from statistical artifact in trait trade-off inference
G. Bullard and D. Jackson
Full code

Run one section at a time (sections denoted by ~~~~~~~~), clearing your environment for each one.

Figure 1, Figure 2, Figure S1
```{r}
rm(list = ls())
#P1: Wide temp range, identifiable slope
#P2: Narrow temp window highlighted; many equally plausible lines
#P3: Ridge in (a,B) from narrow-window bootstrap fits
#P4: Same ridge mapped to (z, sCTmax)

library(tidyverse)
library(patchwork)
library(grid)
library(cowplot)
library(ggbrace)
library(smatr)

set.seed(2)

# helper transforms
ab_to_zs <- function(a, B) {
  # from y = a + B*T to y = (sCTmax - T)/z
  # B = -1/z, a = sCTmax/z
  z <- -1 / B
  sCTmax <- -a / B
  list(z = z, sCTmax = sCTmax)
}

# simulate one underlying "true" TDT line + noise
a_true <- 40       # intercept
B_true <- -1       # slope
sigma_y <- 1    # vertical noise SD for log10(tf)

n_total <- 450

T_full <- runif(n_total, 29, 42)
y_full <- a_true + B_true * T_full + rnorm(n_total, 0, sigma_y)

dat_full <- tibble(Temp = T_full, log_tf = y_full)

# highlight region for narrow temp range
t_lo <- 37
t_hi <- 40
dat_full <- dat_full %>%
  mutate(in_window = Temp >= t_lo & Temp <= t_hi)

dat_window <- dat_full %>% filter(in_window)

# candidate lines that "fit similarly" in the window:

# bootstrap fits from the narrow window
B_lines <- 60

boot_fits <- map_dfr(1:B_lines, function(i) {
  idx <- sample(seq_len(nrow(dat_window)), replace = TRUE)
  fit <- lm(log_tf ~ Temp, data = dat_window[idx, ])
  tibble(
    a = coef(fit)[1],
    B = coef(fit)[2])
})

# transform to z / sCTmax
boot_fits <- boot_fits %>%
  mutate(
    z = -1 / B,
    sCTmax = -a / B
  ) %>%
  filter(is.finite(z), is.finite(sCTmax), z > 0)

# also compute the "true" z, sCTmax for reference
zs_true <- ab_to_zs(a_true, B_true)
z_true <- zs_true$z
s_true <- zs_true$sCTmax


# Panel 1: wide range (single best fit)
fit_full <- lm(log_tf ~ Temp, data = dat_full)

p1 <- ggplot(dat_full, aes(Temp, log_tf)) +
  geom_point(size = 1.7, alpha = 0.85) +
  geom_abline(intercept = coef(fit_full)[1], slope = coef(fit_full)[2], linewidth = 1.3,color="darkorange") +
  labs(
    title = "",
    x = "Temperature (°C)",
    y = expression(log[10]*"(time to failure)")
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())

# Panel 2: narrow window highlighted + many plausible lines
# Build lines for plotting in panel 2
t_grid <- seq(min(dat_full$Temp), max(dat_full$Temp), length.out = 100)

lines_df <- boot_fits[1:min(60, nrow(boot_fits)), ] %>%
  mutate(draw = row_number()) %>%
  tidyr::expand_grid(Temp = t_grid) %>%
  mutate(log_tf = a + B * Temp)

p2 <- ggplot(dat_full, aes(Temp, log_tf)) +
  geom_point(data = dat_full %>% filter(!in_window), color = "grey70", size = 1.5, alpha = 0.6) +
  geom_point(data = dat_full %>% filter(in_window),  size = 1.8, alpha = 0.9) +
  annotate("rect", xmin = t_lo, xmax = t_hi, ymin = -Inf, ymax = Inf,
           alpha = 0.08, fill = "grey10") +
  geom_line(data = lines_df, aes(Temp, log_tf, group = draw),
            alpha = 0.18, linewidth = 0.7,color="pink") +
  geom_smooth(data = dat_full, method="lm",
            alpha = 0.18, linewidth = 1.3,color="hotpink",fullrange=TRUE,se=FALSE) +
  labs(
    title = "",
    x = "Temperature (°C)",
    y = expression(log[10]*"(time to failure)")
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())+
  ylab("")

# Panel 3: ridge in (a, B)
p3 <- ggplot(boot_fits, aes(a, B)) +
  geom_point(
    size = 2,
    alpha = 0.75,
    position = position_jitter(width = 0.02, height = 0.02)
  ) +
  geom_point(aes(x = a_true, y = B_true), shape = 4, stroke = 1.2, size = 4, color="hotpink") +
  labs(
    title = "",
    x = "y-Intercept (\U03B1)",
    y = "Slope (\u03B2)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank()) +
  xlab("") 

#this one is x-intercepts

# Panel 4: mapped ridge in (z, sCTmax)
p4 <- ggplot(boot_fits, aes(sCTmax, z)) +
  geom_point(size = 2, alpha = 0.75) +
  geom_point(aes(x = s_true, y = z_true), shape = 4, stroke = 1.2, size = 4, color="hotpink") +
  labs(title = "", x = expression(sCT[max]),y = "Thermal Senstitivity (z)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank())+
  xlab("")+
  stat_ellipse(color = "grey70", linewidth = 0.8)



```

```{r}
p1.f0 = p1+
  ylab("Response")+
  xlab("Predictor")+
  theme(axis.text=element_blank())
p2.f0 = p2 +
  ylab("Response")+
  xlab("Predictor")+
  theme(axis.text=element_blank())
p3.f0 = ggplot(boot_fits, aes(a, B)) +
  geom_point(size = 2,alpha = 0.75,position = position_jitter(width = 0.02, height = 0.02))+
  geom_point(aes(x = a_true, y = B_true), shape = 4, stroke = 1.2, size = 4, color="hotpink")+
  labs(title= "",x="y-Intercept (\U03B1)",y="Slope (\u03B2)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(),axis.text=element_blank())+
  labs(tag="C")

#x-intercept
x.int.true = -a_true/B_true

p4.f0 = ggplot(boot_fits, aes(sCTmax, B)) +
  geom_point(size = 2,alpha = 0.75,position = position_jitter(width = 0.02, height = 0.02))+
  geom_point(aes(x = x.int.true, y = B_true), shape = 4, stroke = 1.2, size = 4, color="hotpink")+
  labs(title= "",x="x-Intercept",y="Slope (\u03B2)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(),axis.text=element_blank())+
  labs(tag="D")

# Figure 1
set.seed(12)
ggdraw() +
  draw_plot((p1.f0 | p2.f0) / (p3.f0 | p4.f0)) +
  draw_label("Restricted sampling + sampling error = spurious correlation",               
             x = 0.48, y = 0.92,   
             fontface = "italic", size = 12) +
  draw_label("True Parameter", 
             x = 0.65, y = 0.34, 
             fontface = "italic", size = 12, color = "hotpink") +
    draw_label("True Parameter", 
             x = 0.28, y = 0.34, 
             fontface = "italic", size = 12, color = "hotpink") +
  draw_label("Observed Temperature Range", 
             x = 0.83, y = 0.86, 
             fontface = "italic", size = 9, color = "gray40")
```

```{r}
# Bottom row: hierarchical case (multiple lines)

# simulate multiple true lines (same slope, different intercepts)
set.seed(3)
line_colors <- c("#7B3294", "#E69F00", "#0072B2", "#009E73")
light_line_colors <- colorspace::lighten(line_colors, 0.5)
n_lines <- length(line_colors)

a_lines <- a_true + c(-4, -1.5, 1.5, 4)  # spread intercepts for clarity
B_lines_true <- rep(B_true, n_lines)

n_per_line <- 35

multi_dat <- map_dfr(1:n_lines, function(i) {

  Temp <- runif(n_per_line, 29, 41)
  log_tf <- a_lines[i] + B_lines_true[i] * Temp + rnorm(n_per_line, 0, sigma_y)

  tibble(
    Temp = Temp,
    log_tf = log_tf,
    line = paste0("L", i))})

multi_dat <- multi_dat %>%
  mutate(in_window = Temp >= t_lo & Temp <= t_hi)

# Panel 5: multiple true lines (full range)
p5 <- ggplot(multi_dat, aes(Temp, log_tf, color = line)) +
  geom_point(size = 1.7, alpha = 0.85) +
  geom_abline(
    data = tibble(a = a_lines, B = B_lines_true, line = paste0("L",1:n_lines)),
    aes(intercept = a, slope = B, color = line),
    linewidth = 1.2,
    show.legend = FALSE) +
  scale_color_manual(values = line_colors) +
  labs(
    title = "",
    x = "Temperature (°C)",
    y = expression(log[10]*"(time to failure)")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "none")+
  ylim(-5,18)


# bootstrap fits per line using restricted window
boot_multi <- map_dfr(unique(multi_dat$line), function(L) {

  dat_line <- multi_dat %>% filter(line == L, in_window)

  map_dfr(1:25, function(i) {

    idx <- sample(seq_len(nrow(dat_line)), replace = TRUE)
    fit <- lm(log_tf ~ Temp, data = dat_line[idx, ])

    tibble(
      line = L,
      a = coef(fit)[1],
      B = coef(fit)[2])})})

boot_multi <- boot_multi %>%
  mutate(
    z = -1 / B,
    sCTmax = -a / B
  ) %>%
  filter(is.finite(z), is.finite(sCTmax), z > 0)

# candidate lines for P6
t_grid <- seq(29, 41, length.out = 100)

line_df <- boot_multi %>%
  group_by(line) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(draw = row_number()) %>%
  expand_grid(Temp = t_grid) %>%
  mutate(log_tf = a + B * Temp)

# best-fit lines per line
best_fits <- multi_dat %>%
  filter(in_window) %>%
  group_by(line) %>%
  do({
    fit <- lm(log_tf ~ Temp, data = .)
    tibble(a = coef(fit)[1], B = coef(fit)[2])
  })


# Panel 6: restricted window + candidate fits
p6 <- ggplot(multi_dat, aes(Temp, log_tf)) +
  geom_point(data = multi_dat %>% filter(!in_window),
             color = "grey70", size = 1.3, alpha = 0.6) +
  geom_point(data = multi_dat %>% filter(in_window),aes(color = line), size = 1.8, alpha = 1.1) +
  geom_point(data = multi_dat %>% filter(!in_window),aes(color = line), size = 1.8, alpha = 0.25) +
  annotate("rect", xmin = t_lo, xmax = t_hi, ymin = -Inf, ymax = Inf,
           alpha = 0.08, fill = "grey10") +
  geom_line(
    data = line_df,
    aes(Temp, log_tf, group = interaction(line, draw), color = line),
    alpha = 0.15,
    linewidth = 0.7
  ) +
  geom_abline(
    data = best_fits,
    aes(intercept = a, slope = B, color = line),
    linewidth = 1.1
  ) +
  scale_color_manual(values = line_colors) +
  labs(
    title = "",
    x = "Temperature (°C)",
    y = expression(log[10]*"(time to failure)")
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "none")+
  ylim(-5,18)+
  ylab("")


# Panel 7: (a,B) ridge across lines
p7 <- ggplot(boot_multi, aes(a, B, color = line)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = line_colors) +
  labs(
    title = "",
    x = "Intercept (\u03B1)",
    y = "Slope (\u03B2)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "none")+
  ylim(-3,0)+xlim(0,105)

# Panel 8: transformed tradeoff
p8 <- ggplot(boot_multi, aes(sCTmax, z, color = line)) +
  geom_point(size = 2, alpha = 0.8) +
  stat_ellipse(color = "grey70", linewidth = 0.8) +
  scale_color_manual(values = line_colors) +
  labs(
    title = "",
    x = expression(sCT[max]),
    y = "Thermal Sensitivity (z)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "none")+
  ylim(0.25,1.5)+xlim(35,48)

# Combine with original panels

p1 <- p1 + labs(tag = "A")+theme(axis.line = element_line(linewidth = 0.7, colour = "black"))
p2 <- p2 + labs(tag = "B")+theme(axis.line = element_line(linewidth = 0.7, colour = "black"))
p3 <- p3 + labs(tag = "C")+theme(axis.line = element_line(linewidth = 0.7, colour = "black"))
p4 <- p4 + labs(tag = "D")+theme(axis.line = element_line(linewidth = 0.7, colour = "black"))
p5 <- p5 + labs(tag = "E")+theme(axis.line = element_line(linewidth = 0.7, colour = "black"))
p6 <- p6 + labs(tag = "F")+theme(axis.line = element_line(linewidth = 0.7, colour = "black"))
p7 <- p7 + labs(tag = "G")+theme(axis.line = element_line(linewidth = 0.7, colour = "black"))
p8 <- p8 + labs(tag = "H")+theme(axis.line = element_line(linewidth = 0.7, colour = "black"))

arrow_plot <- ggplot() +
  geom_segment(
    aes(x = 0, y = 0.5, xend = 1, yend = 0.5),
    arrow = arrow(length = unit(0.15, "inches")),
    linewidth = 1.2) +
  theme_void() +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))

arrow_plot_vertical <- ggplot() +
  geom_segment(
    aes(x = 0.5, y = 0, xend = 0.5, yend = 1),
    arrow = arrow(
      length = unit(0.15, "inches"),
      ends = "both",
      type = "open"),
    linewidth = 1.2) +
  theme_void() +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))


row_label1 <- ggplot() +
  annotate(
    "text",
    x = 0.5,
    y = 0.5,
    label = "Single Line",
    angle = 90,
    size = 6,
    fontface = "bold"
  ) +
  theme_void() +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1), expand = FALSE) +
  theme(plot.margin = margin(0,0,0,0))

row_label2 <- ggplot() +
  annotate(
    "text",
    x = 0.5,
    y = 0.5,
    label = "Multiple Lines",
    angle = 90,
    size = 6,
    fontface = "bold"
  ) +
  theme_void() +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1), expand = FALSE) +
  theme(plot.margin = margin(0,0,0,0))

label_row <- ggplot() +
  annotate("text", x = 0.28, y = 0.5, label = "Data Space",
           size = 5, fontface = "bold") +
  annotate("text", x = 0.78, y = 0.5, label = "Parameter Space",
           size = 5, fontface = "bold") +
  theme_void() +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1))

brace_row <- ggplot() +
  stat_brace(aes(x = c(0.05, 0.52), y = 0),
             rotate = 0, width = 0.2, outerstart = 0.3) +
  stat_brace(aes(x = c(0.56, 1.02), y = 0),
             rotate = 0, width = 0.2, outerstart = 0.3) +
  theme_void() +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1))

# Stack: label / brace / plots
top_row    <-  label_row
brace_row_ <- plot_spacer() | brace_row

row_single <- row_label1 | (p1 | p2 | p3 | p4)
row_single <- row_single + plot_layout(widths = c(0.08, 1))

row_multi  <- row_label2 | (p5 | p6 | p7 | p8)
row_multi  <- row_multi  + plot_layout(widths = c(0.08, 1))

# Stack rows, control heights only at top level
fig_full <- (top_row / brace_row  / row_single / row_multi) +
  plot_layout(heights = c(0.12, 0.12, 1, 1)) &
  theme(
    axis.text  = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank())

fig_full_final=fig_full +
  inset_element(arrow_plot, 0.28, 2.9, 0.32, 0.78) +
  inset_element(arrow_plot, 0.49, 2.9, 0.53, 0.78) +
  inset_element(arrow_plot, 0.71, 2.9, 0.75, 0.78) +
  inset_element(arrow_plot, 0.28, 0.6, 0.32, 0.78) +
  inset_element(arrow_plot, 0.49, 0.6, 0.53, 0.78) +
  inset_element(arrow_plot, 0.71, 0.6, 0.75, 0.78) +
  inset_element(arrow_plot_vertical, 0, 1.175, 1.25, 1)+
  inset_element(arrow_plot_vertical, 0, 1.175, 1.75, 1)

# Figure 2
ggdraw(fig_full_final) +
  draw_label("Restricted Sampling",               
             x = 0.305, y = 0.86,   # adjust y to sit just above the brace
             fontface = "italic", size = 12) +
  draw_label("Non-Linear Reparameterization", 
             x = 0.769, y = 0.855, 
             fontface = "italic", size = 12)+
  draw_label("True Parameter", 
             x = 0.685, y = 0.77, 
             fontface = "italic", size = 12, color="hotpink")+
  draw_label("Observed Temperature Range", 
             x = 0.47, y = 0.86, 
             fontface = "italic", size = 9, color="gray40")

```

```{r}
#Figure S1
#bottom row Figure 2 but with variable slopes
#simulate multiple generating lines (different slopes, different intercepts)

set.seed(3)
line_colors = c("#7B3294", "#E69F00", "#0072B2", "#009E73")
light_line_colors <- colorspace::lighten(line_colors, 0.5)
n_lines = length(line_colors)

a_lines = a_true + c(-4, -1.5, 1.5, 4)  
B_lines_true <- B_true + runif(4,-0.3,0.3)

n_per_line <- 35

multi_dat <- map_dfr(1:n_lines, function(i) {

  Temp <- runif(n_per_line, 29, 41)
  log_tf <- a_lines[i] + B_lines_true[i] * Temp + rnorm(n_per_line, 0, sigma_y)

  tibble(
    Temp = Temp,
    log_tf = log_tf,
    line = paste0("L", i))})

multi_dat <- map_dfr(1:n_lines, function(i) {

  Temp <- runif(n_per_line, 29, 41)
  log_tf <- a_lines[i] + B_lines_true[i] * Temp + rnorm(n_per_line, 0, sigma_y)

  tibble(
    Temp = Temp,
    log_tf = log_tf,
    line = paste0("L", i))})

multi_dat = multi_dat %>%
  mutate(in_window = Temp >= t_lo & Temp <= t_hi)

#panel 5 Figure 2
p5s = ggplot(multi_dat, aes(Temp, log_tf, color = line)) +
  geom_point(size = 1.7, alpha = 0.85) +
  geom_abline(
    data = tibble(a = a_lines, B = B_lines_true, line = paste0("L",1:n_lines)),
    aes(intercept = a, slope = B, color = line),
    linewidth = 1.2,
    show.legend = FALSE) +
  scale_color_manual(values = line_colors) +
  labs(
    title = "",
    x = "Temperature (°C)",
    y = expression(log[10]*"(time to failure)")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "none")+
  ylim(-5,18)+
  labs(tag="A")

# bootstrap fits per line using restricted window
boot_multi = map_dfr(unique(multi_dat$line), function(L) {

  dat_line = multi_dat %>% filter(line == L, in_window)

  map_dfr(1:25, function(i) {

    idx = sample(seq_len(nrow(dat_line)), replace = TRUE)
    fit = lm(log_tf ~ Temp, data = dat_line[idx, ])

    tibble(
      line = L,
      a = coef(fit)[1],
      B = coef(fit)[2])})})

boot_multi = boot_multi %>%
  mutate(z = -1 / B, sCTmax = -a / B) %>%
  filter(is.finite(z), is.finite(sCTmax), z > 0)

#candidate lines panel 6
t_grid = seq(29, 41, length.out = 100)

line_df = boot_multi %>%
  group_by(line) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(draw = row_number()) %>%
  expand_grid(Temp = t_grid) %>%
  mutate(log_tf = a + B * Temp)

# best-fit lines per group

best_fits <- multi_dat %>%
  filter(in_window) %>%
  group_by(line) %>%
  do({
    fit <- lm(log_tf ~ Temp, data = .)
    tibble(a = coef(fit)[1], B = coef(fit)[2])
  })

#panel 6
p6s = ggplot(multi_dat, aes(Temp, log_tf)) +
  geom_point(data = multi_dat %>% filter(!in_window),color = "grey70", size = 1.3, alpha = 0.6) +
  geom_point(data = multi_dat %>% filter(in_window),aes(color = line), size = 1.8, alpha = 1.1) +
  geom_point(data = multi_dat %>% filter(!in_window),aes(color = line), size = 1.8, alpha = 0.25) +
  annotate("rect", xmin = t_lo, xmax = t_hi, ymin = -Inf, ymax = Inf, alpha = 0.08, fill = "grey10") +
  geom_line(data = line_df, aes(Temp, log_tf, group = interaction(line, draw), color = line), alpha = 0.15, linewidth = 0.7) +
  geom_abline(data = best_fits, aes(intercept = a, slope = B, color = line), linewidth = 1.1) +
  scale_color_manual(values = line_colors) +
  labs(title = "", x = "Temperature (°C)", y = expression(log[10]*"(time to failure)")) +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "none")+
  ylim(-5,18)+
  ylab("")+
  labs(tag="B")

#panel 7
p7s = ggplot(boot_multi, aes(a, B, color = line)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_manual(values = line_colors) +
  labs(title = "", x = "Intercept (\u03b1)", y = "Slope (\u03b2)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "none")+
  ylim(-2.5,0)+xlim(0,105)+
  labs(tag="C")

#panel 8
p8s = ggplot(boot_multi, aes(sCTmax, z, color = line)) +
  geom_point(size = 2, alpha = 0.8) +
  stat_ellipse(color = "grey70", linewidth = 0.8) +
  scale_color_manual(values = line_colors) +
  labs(title = "", x = expression(sCT[max]), y = "Thermal Sensitivity (z)") +
  theme_minimal(base_size = 12) +
  theme(panel.grid = element_blank(), legend.position = "none")+
  ylim(0.25,1.5)+xlim(37,40)+
  labs(tag="D")
#Figure S1
full.fig  = (p5s | p6s | p7s | p8s)
full.fig
```

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

Figure 3

```{r}
rm(list = ls())
#given that this figure is a qualitative proof of concept, not a quantitative analysis, I've included these scalars to allow you to adjust the variance in the error parameters. The output is extremely sensitive to these parameters, but given the unidentifiability of the model, they are hard to estimate (significant uncertainty in our estimates), so we scale them to show that there IS an uncorrelated distribution that qualitatively resembles the Soto et al data after resampling. Not that we have a way of accurately identifying that distribution, since B is likely skewed, not normal. But these parameters are a place to start. 

#Note the codependence between these parameters. Increasing e too much will result in an error because it will force B to be 0

#These values were obtained via optimization (code below Figure 3)
B_var_scalar=0.1
a_var_scalar=0.04375
e_var_scalar=1.3

library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(MASS)
library(scales)
library(cowplot)

set.seed(1)

#Settings
file_path   = "Soto_etal_data/KO_4temp_100DGRP.csv"  # change if needed
sex_label   = "F"                     # "F" or "M"
T0          = 38
temps_sim   = 37:40
n_per_temp  = 5
n_show_lines = 10                     # for panel C

# Helper functions
zs_to_ab = function(z, s) {
  B = -1 / z
  a = s / z
  list(a = a, B = B)
}

ab_to_zs = function(a, B) {
  z = -1 / B
  s = -a / B
  list(z = z, sCTmax = s)
}

#generate a dataset similar to Soto et al, but assuming uncorrelated slope and intercept.
simulate_one_dataset_ab = function(mu_a, mu_B, sigma_a, sigma_B, sigma_e,
                                    n_lines = 100, temps = 37:40, n_per_temp = 5) {

  # draw "true" parameters 
  a_true = rnorm(n_lines, mean = mu_a, sd = sigma_a)
  B_true = rnorm(n_lines, mean = mu_B, sd = sigma_B)

  # build dataset
  sim_dat = tidyr::expand_grid(
    line_sim = seq_len(n_lines),
    Temp = temps,
    rep = seq_len(n_per_temp)
  )

  # assign outside mutate to avoid scoping issues
  sim_dat$a = a_true[sim_dat$line_sim]
  sim_dat$B = B_true[sim_dat$line_sim]

  # simulate on log scale
  sim_dat$mu = sim_dat$a + sim_dat$B * sim_dat$Temp
  sim_dat$y  = rnorm(nrow(sim_dat), mean = sim_dat$mu, sd = sigma_e)

  # average on log scale
  fits = sim_dat %>%
    dplyr::group_by(line_sim, Temp) %>%
    dplyr::summarise(y = mean(y), .groups = "drop") %>%
    dplyr::group_by(line_sim) %>%
    dplyr::summarise(
      a_hat = coef(lm(y ~ Temp))[1],
      B_hat = coef(lm(y ~ Temp))[2],
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      z_hat = -1 / B_hat,
      sCTmax_hat = -a_hat / B_hat
    )

  tibble_true = tibble::tibble(
    line_sim = seq_len(n_lines),
    a = a_true,
    B = B_true
  )

  list(
    true_pars = tibble_true,
    sim_dat = sim_dat,
    recovered = fits
  )
}

#display a few of the lines in TDT space to demonstrate the concept
pick_points_for_panelC = function(dat, n_per_temp = 5) {
  dat %>%
    dplyr::group_by(line_sim, Temp) %>%
    dplyr::slice_sample(n = n_per_temp, replace = TRUE) %>%
    dplyr::ungroup()
}

#density plot for z/sCTmax parameterization
dens_zs = function(df, title = "", subtitle = "", xlims, ylims) {
  ggplot(df, aes(x = sCTmax, y = z)) +
    stat_density_2d(aes(fill = after_stat(level)), geom = "polygon") +
    scale_fill_viridis_c(trans = "log", guide = "none") +
    coord_cartesian(xlim = xlims, ylim = ylims) +
    labs(
      title = title,
      subtitle = subtitle,
      x = expression(sCT[max]),
      y = "z"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold")
    )
}

#density plot for a/B parameterization
dens_ab = function(df, x_col = "a", y_col = "B", title = "", subtitle = "", xlims, ylims) {
  ggplot(df, aes(x = .data[[x_col]], y = .data[[y_col]])) +
    stat_density_2d(aes(fill = after_stat(level)), geom = "polygon") +
    scale_fill_viridis_c(trans = "log", guide = "none") +
    coord_cartesian(xlim = xlims, ylim = ylims) +
    labs(
      title = title,
      subtitle = subtitle,
      x = "Intercept (\u03b1)",
      y = "Slope (\u03b2)"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid = element_blank(),
      plot.title = element_text(face = "bold")
    )
}

# Read + Format Soto et al data
# -------------------------
soto.dat = read.csv(file_path)

soto.dat$time   = soto.dat$KO / 60
soto.dat$Temp   = soto.dat$temp
soto.dat$line   = soto.dat$Geno

# Empirical LSE table
real.pars.soto = data.frame(
  z = numeric(),
  sCTmax = numeric(),
  a = numeric(),
  B = numeric(),
  sex = character(),
  line = numeric(),
  stringsAsFactors = FALSE
)

for (geno_i in unique(soto.dat$Geno)) {
  data_i = soto.dat[soto.dat$Geno == geno_i, ]

  for (sex_j in c("F", "M")) {
    data_ij = data_i[data_i$sex == sex_j, ]
    if (nrow(data_ij) == 0) next
    if (length(unique(data_ij$Temp)) < 2) next

    data_means_ij = data_ij %>%
      group_by(Temp) %>%
      summarise(mean_tf = mean(time, na.rm = TRUE), .groups = "drop")

    if (nrow(data_means_ij) < 2) next

    lm_ij = lm(log10(mean_tf) ~ Temp, data = data_means_ij)
    a_hat = coef(lm_ij)[1]
    B_hat = coef(lm_ij)[2]

    real.pars.soto = rbind(
      real.pars.soto,
      data.frame(
        z = -1 / B_hat,
        sCTmax = -a_hat / B_hat,
        a = a_hat,
        B = B_hat,
        sex = sex_j,
        line = geno_i,
        stringsAsFactors = FALSE
      )
    )
  }
}
rownames(real.pars.soto) = NULL

# restrict to one sex
dat_sex = soto.dat %>% filter(sex == sex_label)
pars_sex = real.pars.soto %>%
  filter(sex == sex_label) %>%
  filter(is.finite(z), is.finite(sCTmax), is.finite(a), is.finite(B))

# Null calibration 
# middle variance setting in a,B
a_obs = real.pars.soto$a[real.pars.soto$sex==sex_label]
B_obs = real.pars.soto$B[real.pars.soto$sex==sex_label]

# Full multi-step parameter estimation
temps = sort(unique(dat_sex$Temp))
T_bar = mean(temps)
SS_Temp = sum((temps - T_bar)^2)
n_temps = length(temps)

# estimate sigma_e^2 using means per temp 
estimate_sigma_e2_full = function(dat_sex, n_per_temp) {
  fits = dat_sex %>%
    group_by(line, Temp) %>%
    summarise(y = mean(log10(time)), .groups = "drop") %>%
    group_by(line) %>%
    group_modify(~{
      if (nrow(.x) < 2) return(NULL)
      fit = lm(y ~ Temp, data = .x)
      tibble(sigma2 = summary(fit)$sigma^2)
    })
  
  median(fits$sigma2, na.rm = TRUE) * n_per_temp
}

sigma_e2 = estimate_sigma_e2_full(dat_sex, n_per_temp)

# slope variance
var_B_obs = var(B_obs, na.rm = TRUE)
sigma_B2 = var_B_obs - sigma_e2 / (n_per_temp * SS_Temp)
sigma_B2 = sigma_B2*B_var_scalar
sigma_B2 = max(sigma_B2, 0)

# intercept variance
var_a_obs = var(a_obs, na.rm = TRUE)

inflation_a = sigma_e2 * (
  1 / (n_per_temp * n_temps) +
  (T_bar^2) / (n_per_temp * SS_Temp)
)

sigma_a2 = var_a_obs - inflation_a
# scale a
sigma_a2 = max(sigma_a2, 0)* a_var_scalar

sigma_e2 = sigma_e2*e_var_scalar

# covariance
cov_obs = cov(a_obs, B_obs, use = "complete.obs")
cov_aB_true = cov_obs + sigma_e2 * T_bar / (n_per_temp * SS_Temp)

# final parameters
sigma_a = sqrt(sigma_a2)
sigma_B = sqrt(sigma_B2)
sigma_e = sqrt(sigma_e2)

# null generating distribution (pA)
mu_a = mean(a_obs, na.rm = TRUE)
mu_B = mean(B_obs, na.rm = TRUE)

#null distribution for pA
null_ab = MASS::mvrnorm(
  n = 8000,
  mu = c(mu_a, mu_B),
  Sigma = matrix(
    c(sigma_a2, 0,
      0, sigma_B2),
    2, 2
  )
) %>%
  as.data.frame()
colnames(null_ab) = c("a", "B")
sigma_e = sqrt(sigma_e2)
n_lines_this = nrow(pars_sex)


# Simulate one dataset under null in a,B
sim_out = simulate_one_dataset_ab(
  mu_a = mu_a,
  mu_B = mu_B,
  sigma_a = sigma_a,
  sigma_B = sigma_B,
  sigma_e = sigma_e,
  n_lines = n_lines_this,
  temps = temps_sim,
  n_per_temp = n_per_temp
)

true_ab_sim = sim_out$true_pars
sim_dat = sim_out$sim_dat
recovered_sim = sim_out$recovered

# pF/G: Empirical z,sCTmax
emp_zs = pars_sex %>%
  dplyr::select(z, sCTmax) %>%
  filter(is.finite(z), is.finite(sCTmax), z > 0)

#pC: simulated TDT curves
#5 points per line per temp
#demonstrates the sampling process 

n_show_lines = min(n_show_lines, n_lines_this)
show_lines = sample(true_ab_sim$line_sim, n_show_lines)

sim_dat_show = sim_dat %>%
  filter(line_sim %in% show_lines)

sim_dat_show5 = pick_points_for_panelC(sim_dat_show, n_per_temp = 5)

true_show = true_ab_sim %>%
  filter(line_sim %in% show_lines)

line_cols = scales::hue_pal()(n_show_lines)
names(line_cols) = as.character(show_lines)

# recovered simulated z,sCTmax
sim_zs = recovered_sim %>%
  transmute(z = z_hat, sCTmax = sCTmax_hat) %>%
  filter(is.finite(z), is.finite(sCTmax), z > 0)

# Common limits
# z,sCTmax panels: encompass all data across bottom row panels
# a,B gets its own broad range for top row panels

# pD: Recovered a_hat/B_hat from simulation (not true parameters)
recovered_ab_sim = data.frame(
  a = recovered_sim$a_hat,
  B = recovered_sim$B_hat
)

# pF: Empirical Soto a,B estimates — "strong correlation"
emp_ab_df = data.frame(a = a_obs, B = B_obs)

# Shared a,B limits across pA, pD, pF
ab_shared = bind_rows(
  null_ab,
  recovered_ab_sim,
  emp_ab_df
)


zs_all = bind_rows(
  emp_zs %>% mutate(source = "emp"),
  sim_zs %>% mutate(source = "sim")
)

xlims_zs = range(zs_all$sCTmax, na.rm = TRUE)
ylims_zs = range(zs_all$z, na.rm = TRUE)
xpad_zs = 0.04 * diff(xlims_zs)
ypad_zs = 0.04 * diff(ylims_zs)
xlims_zs = c(xlims_zs[1] - xpad_zs, xlims_zs[2] + xpad_zs)
ylims_zs = c(ylims_zs[1] - ypad_zs, ylims_zs[2] + ypad_zs)

ab_all = bind_rows(
  null_ab %>% mutate(source = "null"),
  tibble(a = a_obs, B = B_obs, source = "emp")
)
xlims_ab = range(ab_all$a, na.rm = TRUE)
ylims_ab = range(ab_all$B, na.rm = TRUE)
xpad_ab = 0.04 * diff(xlims_ab)
ypad_ab = 0.04 * diff(ylims_ab)

xlims_ab_shared = range(ab_shared$a, na.rm = TRUE)
ylims_ab_shared = range(ab_shared$B, na.rm = TRUE)
xpad_ab_shared = 0.04 * diff(xlims_ab_shared)
ypad_ab_shared = 0.04 * diff(ylims_ab_shared)
xlims_ab_shared = c(xlims_ab_shared[1] - xpad_ab_shared, xlims_ab_shared[2] + xpad_ab_shared)
ylims_ab_shared = c(ylims_ab_shared[1] - ypad_ab_shared, ylims_ab_shared[2] + ypad_ab_shared)

# correlations for annotations
r_emp = cor(emp_zs$z, emp_zs$sCTmax, use = "complete.obs")
r_sim = cor(sim_zs$z, sim_zs$sCTmax, use = "complete.obs")


# Transform null a,B cloud into z/sCTmax space
null_zs_transformed = ab_to_zs(null_ab$a, null_ab$B) %>%
  as.data.frame() %>%
  filter(is.finite(z), is.finite(sCTmax), z > 0)

# Limits for the transformed null cloud
xlims_null_zs = range(null_zs_transformed$sCTmax, na.rm = TRUE)
ylims_null_zs = range(null_zs_transformed$z, na.rm = TRUE)
xpad_null = 0.04 * diff(xlims_null_zs)
ypad_null = 0.04 * diff(ylims_null_zs)
xlims_null_zs = c(xlims_null_zs[1] - xpad_null, xlims_null_zs[2] + xpad_null)
ylims_null_zs = c(ylims_null_zs[1] - ypad_null, ylims_null_zs[2] + ypad_null)

# annotation layer
annot_label = function(label) {
  geom_text(
    data = data.frame(x = -Inf, y = Inf, label = label),
    aes(x = x, y = y, label = label),
    hjust = -0.1, vjust = 1.5, size = 4, inherit.aes = FALSE
  )
}

arrow_plot <- ggplot() +
  geom_segment(
    aes(x = 0, y = 0.5, xend = 1, yend = 0.5),
    arrow = arrow(length = unit(0.15, "inches")),
    linewidth = 1.2
  ) +
  theme_void() +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))

double_arrow_plot <- ggplot() +
  geom_segment(
    aes(x = 0, y = 0.5, xend = 1, yend = 0.5),
    arrow = arrow(length = unit(0.15, "inches"),ends="both"),
    linewidth = 1.2,
  ) +
  theme_void() +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))

arrow_plot_vertical <- ggplot() +
  geom_segment(
    aes(x = 0.5, y = 0, xend = 0.5, yend = 1),
    arrow = arrow(length = unit(0.15, "inches"), ends="both"), linewidth = 1.2) +
  theme_void() +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))

dashed_plot_vertical <- ggplot() +
  geom_segment(
    aes(x = 0.5, y = 0, xend = 0.5, yend = 1),
    linetype = "dashed", linewidth = 1.2) +
  theme_void() +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))
```

```{r}
# BUILD PANELS (Figure 3)
# SMA helper

sma_fit <- function(x, y){
  r <- cor(x, y, use="complete.obs")
  slope <- sign(r) * (sd(y, na.rm=TRUE) / sd(x, na.rm=TRUE))
  intercept <- mean(y, na.rm=TRUE) - slope * mean(x, na.rm=TRUE)
  list(
    slope = slope,
    intercept = intercept,
    rho = r
  )
}

# fits for each panel
fit_B <- sma_fit(null_zs_transformed$sCTmax, null_zs_transformed$z)
fit_C <- sma_fit(recovered_ab_sim$a, recovered_ab_sim$B)
fit_F <- sma_fit(sim_zs$sCTmax, sim_zs$z)
fit_D <- sma_fit(emp_ab_df$a, emp_ab_df$B)
fit_G <- sma_fit(emp_zs$sCTmax, emp_zs$z)

beta_lab <- function(b){
  paste0("\u03B2 = ", sprintf("%.2f", b))
}

rho_lab <- function(r){
  paste0("\u03C1 = ", sprintf("%.2f", r))
}

# pA

pA = dens_ab(
  null_ab,
  x_col = "a", y_col = "B",
  title = "", subtitle = "",
  xlims = range(null_ab$a),
  ylims = range(null_ab$B)
) +
  annot_label("no covariance") +
  coord_cartesian(xlim = xlims_ab_shared, ylim = ylims_ab_shared, clip = "off")+
  labs(tag = "A")+
  theme(axis.line = element_line(linewidth = 0.7, colour = "black"))+
  labs(x="Intercept (\u03b1)", y="Slope (\u03b2)")

# pE (null → zs)

pE = dens_zs(
  null_zs_transformed,
  title = "", subtitle = "",
  xlims = xlims_null_zs,
  ylims = ylims_null_zs
) +
  geom_abline(
    intercept = fit_B$intercept,
    slope = fit_B$slope,
    color = "#9C27FF",
    linewidth = 1.2,
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = 41.5, y = 5.5,
    label = beta_lab(fit_B$slope),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 41.5, y = 5.3,
    label = rho_lab(fit_B$rho),
    hjust = 0
  ) +
  annot_label("high induced covariance") +
  coord_cartesian(xlim = c(41,49), ylim = c(3,6), clip = "off")+
  labs(tag = "E")+
  draw_label("Nonlinear Reparameterization", x = 49.5, y = 6.65, fontface = "italic", size = 9)+
  theme(axis.line = element_line(linewidth = 0.7, colour = "black"))

# pB (TDT curves)

pB = ggplot() +
  geom_point(
    data = sim_dat_show5,
    aes(x = Temp, y = y, color = factor(line_sim)),
    alpha = 0.5, size = 1.5
  ) +
  geom_abline(
    data = true_show,
    aes(intercept = a, slope = B, color = factor(line_sim)),
    linewidth = 1.0, show.legend = FALSE
  ) +
  scale_color_manual(values = line_cols) +
  labs(
    tag = "B", subtitle = "",
    x = "Temperature (°C)",
    y = expression(log[10]*"(time to failure)")
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )+
  coord_cartesian(clip = "off")+
  draw_label("Reestimate linear fits \nbased on generated points", x = 39.5, y = 2.52, fontface = "italic", size = 9)+
  theme(axis.line = element_line(linewidth = 0.7, colour = "black"))

# pC (recovered a,B)

pC = dens_ab(
  recovered_ab_sim,
  x_col = "a", y_col = "B",
  title = "Simulated Data", subtitle = "     (Under Null)",
  xlims = range(recovered_ab_sim$a),
  ylims = range(recovered_ab_sim$B)
) +
  geom_abline(
    intercept = fit_C$intercept,
    slope = fit_C$slope,
    color = "#9C27FF",
    linewidth = 1.2,
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = 12,
    y = max(recovered_ab_sim$B),
    label = beta_lab(fit_C$slope),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 12,
    y = max(recovered_ab_sim$B) * 1.1,
    label = rho_lab(fit_C$rho),
    hjust = 0
  ) +
  annot_label("high induced covariance") +
  coord_cartesian(xlim = xlims_ab_shared, ylim = ylims_ab_shared, clip = "off")+ 
  labs(tag = "C")+
  theme(axis.line = element_line(linewidth = 0.7, colour = "black"))+
  labs(x="Intercept (\u03b1)", y="Slope (\u03b2)")

# pF (sim zs)

pF = dens_zs(
  sim_zs,
  title = "", subtitle = "",
  xlims = c(41,49),
  ylims = c(2,6)
) +
  geom_abline(
    intercept = fit_F$intercept,
    slope = fit_F$slope,
    color = "#9C27FF",
    linewidth = 1.2,
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = 41.5, y = 5.5,
    label = beta_lab(fit_F$slope),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 41.5, y = 5.3,
    label = rho_lab(fit_F$rho),
    hjust = 0
  ) +
  annot_label("high induced covariance") +
  coord_cartesian(xlim = c(41,49), ylim = c(3,6), clip = "off")+ 
  labs(tag = "F")+
  theme(axis.line = element_line(linewidth = 0.7, colour = "black"))

# pD (empirical a,B)

pD = dens_ab(
  emp_ab_df,
  x_col = "a", y_col = "B",
  title = "Empirical Data", subtitle = "",
  xlims = range(emp_ab_df$a),
  ylims = range(emp_ab_df$B)
) +
  geom_abline(
    intercept = fit_D$intercept,
    slope = fit_D$slope,
    color = "#9C27FF",
    linewidth = 1.2,
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = 11,
    y = max(recovered_ab_sim$B),
    label = beta_lab(fit_D$slope),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 11,
    y = max(recovered_ab_sim$B)*1.1 ,
    label = "\u03C1 < -0.99",
    hjust = 0
  ) +
  annot_label("high induced covariance") +
  coord_cartesian(xlim = xlims_ab_shared, ylim = ylims_ab_shared, clip = "off")+ 
  labs(tag = "D")+
  theme(axis.line = element_line(linewidth = 0.7, colour = "black"))+
  labs(x="Intercept (\u03b1)", y="Slope (\u03b2)")

# pG (empirical zs)

pG = dens_zs(
  emp_zs,
  title = "", subtitle = "",
  xlims = c(43,47.5),
  ylims = c(3,6)
) +
  geom_abline(
    intercept = fit_G$intercept,
    slope = fit_G$slope,
    color = "#9C27FF",
    linewidth = 1.2,
    linetype = "dashed"
  ) +
  annotate(
    "text",
    x = 42.5, y = 5.5,
    label = beta_lab(fit_G$slope),
    hjust = 0
  ) +
  annotate(
    "text",
    x = 42.5, y = 5.3,
    label = rho_lab(fit_G$rho),
    hjust = 0
  ) +
  annot_label("high induced covariance") +
  coord_cartesian(xlim = c(41,49), ylim = c(3,6), clip = "off")+ 
  labs(tag = "G")+
  theme(axis.line = element_line(linewidth = 0.7, colour = "black"))

pad_labels <- function(x, width = 6) {
  formatted <- scales::label_number(accuracy = 0.1)(x)
  formatC(formatted, width = width, flag = " ")  # pad with leading spaces
}
#this is plot B but everything is blank. A scaled plot spacer.
pB_fake = ggplot() +
  geom_point(
    data = sim_dat_show5,
    aes(x = Temp, y = y, color = factor(line_sim)),
    alpha = 0.5, size = 1.5
  ) +
  geom_abline(
    data = true_show,
    aes(intercept = a, slope = B, color = factor(line_sim)),
    linewidth = 1.0, show.legend = FALSE
  ) +
  scale_color_manual(values = line_cols) +
  labs(tag="B",subtitle = "",
    x = "Temperature (°C)",
    y = expression(log[10]*"(time to failure)")
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  )+
  coord_cartesian(clip = "off") +
    scale_color_manual(values = rep("white", 20)) +   # adjust length to n categories
  scale_fill_manual(values = rep("white", 20)) +
  theme(
    axis.text  = element_text(color = "white"),
    axis.title = element_text(color = "white"),
    axis.ticks = element_line(color = "white"),
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background   = element_rect(fill = "white", color = NA),
    legend.position = "none"
  )+theme(
  axis.line = element_line(color = "white"),
  plot.tag = element_text(color="white")
)

# Combine

divider <- ggplot() +
  geom_vline(xintercept = 0.5, linetype = "dashed", linewidth = 0.7) +
  theme_void() +
  coord_cartesian(xlim = c(0,1), ylim = c(0,1))

blank_panel <- ggplot() +theme_void() +labs(x = NULL, y = NULL)
fig3_main = (
  (pA | pB | pC | blank_panel | pD) /
  (pE | pB_fake | pF | blank_panel | pG)
) +
  plot_layout(
    widths = c(1,1,1,0.07,1),
    heights = c(1,1),
    guides="collect",
    axes = "collect",
    axis_titles = "collect"
  )+
  inset_element(arrow_plot, 0.12, 4, 0.17, 0)+
  inset_element(arrow_plot, 0.35, 4.4, 0.40, 0)+
  inset_element(double_arrow_plot, 0.64, 4, 0.76, 0)+
  inset_element(double_arrow_plot, 0.64, 1, 0.76, 0)+
  inset_element(dashed_plot_vertical,0, 2.6, 1.41, -0.2)+
  inset_element(arrow_plot_vertical,0, 1.25, 1.86, 1.05)+
  inset_element(arrow_plot_vertical,0, 1.25, 1.09, 1.05)+
  inset_element(arrow_plot_vertical,0, 1.25, 0.15, 1.05)&
  theme(plot.title = element_text(size = 14, face = "bold"))

fig3_main
```



Optimization to obtain parameters above
```{r}
compute_moments <- function(df) {
  c(mean_a = mean(df$a, na.rm = TRUE),
    mean_B = mean(df$B, na.rm = TRUE),
    var_a  = var(df$a, na.rm = TRUE),
    var_B  = var(df$B, na.rm = TRUE),
    cor = cor(df$a, df$B, use = "complete.obs"))}
  
  moment_loss <- function(sim_mom, emp_mom) {
  sum((sim_mom - emp_mom)^2, na.rm = TRUE)}

# ============================================================
# SETTINGS
# ============================================================

e_grid = seq(0.8, 1.5, length.out = 8)
a_grid = seq(0.01, 0.1, length.out = 9)
B_grid = seq(0.1, 0.5, length.out = 9)

n_reps = 10   

grid = expand.grid(
  e_var_scalar = e_grid,
  a_var_scalar = a_grid,
  B_var_scalar = B_grid
)

results = data.frame()

set.seed(123)

# ============================================================
# GRID SEARCH
# ============================================================

for (i in seq_len(nrow(grid))) {

  e_val = grid$e_var_scalar[i]
  a_val = grid$a_var_scalar[i]
  B_val = grid$B_var_scalar[i]

  losses_rep = numeric(n_reps)

  for (r in 1:n_reps) {

    # --- sigma_e ---
    sigma_e2_tmp = estimate_sigma_e2_full(dat_sex, n_per_temp)

    # --- slope variance ---
    var_B_obs = var(B_obs, na.rm = TRUE)

    # --- intercept variance ---
    var_a_obs = var(a_obs, na.rm = TRUE) * a_val

    inflation_a_tmp = sigma_e2_tmp * (
      1 / (n_per_temp * n_temps) +
      (T_bar^2) / (n_per_temp * SS_Temp)
    )
# --- slope variance ---
var_B_obs = var(B_obs, na.rm = TRUE)
sigma_B2_tmp = var_B_obs - sigma_e2_tmp / (n_per_temp * SS_Temp)
#scale B
sigma_B2_tmp = sigma_B2_tmp * B_val

# --- intercept variance ---
var_a_obs = var(a_obs, na.rm = TRUE) 

inflation_a_tmp = sigma_e2_tmp * (
  1 / (n_per_temp * n_temps) +
  (T_bar^2) / (n_per_temp * SS_Temp)
)
#scale a
sigma_a2_tmp = (var_a_obs - inflation_a_tmp) * a_val
#scale error
sigma_e2_tmp = sigma_e2_tmp*e_val

# --- VALIDITY CHECK ---
if (sigma_a2_tmp <= 0 || sigma_B2_tmp <= 0) {
  losses_rep[r] = NA
  next
}

    # --- sigmas ---
    sigma_a_tmp = sqrt(sigma_a2_tmp)
    sigma_B_tmp = sqrt(sigma_B2_tmp)
    sigma_e_tmp = sqrt(sigma_e2_tmp)

    # --- simulate ---
    sim_out_tmp = simulate_one_dataset_ab(
      mu_a = mu_a,
      mu_B = mu_B,
      sigma_a = sigma_a_tmp,
      sigma_B = sigma_B_tmp,
      sigma_e = sigma_e_tmp,
      n_lines = nrow(pars_sex),
      temps = temps_sim,
      n_per_temp = n_per_temp
    )

    recovered_tmp = sim_out_tmp$recovered

    sim_ab_df = data.frame(
      a = recovered_tmp$a_hat,
      B = recovered_tmp$B_hat
    )

    # --- loss ---
    sim_mom = compute_moments(sim_ab_df)
    emp_mom = compute_moments(emp_ab_df)
    losses_rep[r] = moment_loss(sim_mom, emp_mom)
  }

  # remove invalid reps
valid_losses = losses_rep[is.finite(losses_rep)]

if (length(valid_losses) < max(3, n_reps / 3)) {
  # too many invalid reps → discard this grid point
  loss_mean = NA
  loss_sd = NA
} else {
  loss_mean = mean(valid_losses)
  loss_sd = sd(valid_losses)
}

results = rbind(results, data.frame(
  e_var_scalar = e_val,
  a_var_scalar = a_val,
  B_var_scalar = B_val,
  loss_mean = loss_mean,
  loss_sd = loss_sd,
  n_valid = length(valid_losses)
))
results_clean = results %>%
  filter(is.finite(loss_mean))
}

# BEST PARAMETER SET

best_row = results_clean[which.min(results_clean$loss_mean), ]

cat("Best parameters:\n")
print(best_row)
```

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

Table 2 + GWAS analysis 

```{r}
library(tidyverse)
library(data.table)
```


```{r}
soto.dat = read.csv("Soto_etal_data/KO_4temp_100DGRP.csv")


#format the data for the model
soto.dat$time=soto.dat$KO/60
soto.dat$log_tf=log10(soto.dat$time)
soto.dat$Temp = soto.dat$temp
soto.dat$line=soto.dat$Geno
```

We also include centered estimates. Although SNP significance is mostly unchanged by centering, since it only affects non-biological signal.

```{r}
#first, we center with respect to x to reduce unidentifiability in a/B parameterization (centering around a)
Tbar = mean(soto.dat$Temp, na.rm = TRUE)
soto.dat.centered <- soto.dat %>%
  mutate(
    Temp_raw = Temp,
    Temp = Temp_raw - Tbar
  )

centered.pars.soto <- data.frame(
  z = numeric(),
  sCTmax = numeric(),
  a = numeric(),
  B = numeric(),
  sex = character(),
  line=numeric(),
  stringsAsFactors = FALSE
)

for (geno_i in unique(soto.dat.centered$Geno)) {
  data_i <- soto.dat.centered[soto.dat.centered$Geno == geno_i, ]
  for (sex_j in c("F", "M")) {
    data_ij <- data_i[data_i$sex == sex_j, ]
    data_means_ij = data_ij %>%
  group_by(Temp) %>%
  summarise(
    mean_tf = mean(time, na.rm = TRUE),
    .groups = "drop"
  )
    lm_ij <- lm(log10(mean_tf) ~ Temp, data = data_means_ij)
    a_hat <- coef(lm_ij)[1]
    B_hat <- coef(lm_ij)[2]
    sCTmax_ij <- -a_hat / B_hat
    #version of sCTmax with NEW PARAMETERIZATION
    z_ij <- -1 / B_hat
    centered.pars.soto <- rbind(
      centered.pars.soto,
      data.frame(z = z_ij, sCTmax = sCTmax_ij,a=a_hat, B=B_hat, sex = sex_j,line=geno_i, stringsAsFactors = FALSE))
  }}

pca.centered.males <- prcomp(centered.pars.soto[centered.pars.soto$sex=="M", c("a", "B")], scale. = TRUE)
pca.centered.females <- prcomp(centered.pars.soto[centered.pars.soto$sex=="F", c("a", "B")], scale. = TRUE)



#make phenotype files
DGRP2_PC1_Centered = data.frame(DGRP=c(centered.pars.soto$line[centered.pars.soto$sex=="M"]),
                                Male=c(pca.centered.males$x[,1]),
                                Female=c(pca.centered.females$x[,1]))
DGRP2_z_Centered = data.frame(DGRP=c(centered.pars.soto$line[centered.pars.soto$sex=="M"]),
                              Male=c(centered.pars.soto$z[centered.pars.soto$sex=="M"]),
                              Female=c(centered.pars.soto$z[centered.pars.soto$sex=="F"]))
DGRP2_sCTmax_Centered = data.frame(DGRP=c(centered.pars.soto$line[centered.pars.soto$sex=="M"]),
                                   Male=c(centered.pars.soto$sCTmax[centered.pars.soto$sex=="M"]),
                                   Female=c(centered.pars.soto$sCTmax[centered.pars.soto$sex=="F"]))

summary(pca.centered.males)
summary(pca.centered.females)

# #pc1 describes much less of the variance!!! z and sCTmax might actually explain some data now.
# ggplot(centered.pars.soto.male)+
#   geom_histogram(aes(x=PC1))




```


Now get the normal uncentered parameter estimates 
```{r}
real.pars.soto = data.frame(
  z = numeric(),
  sCTmax = numeric(),
  a = numeric(),
  B = numeric(),
  sex = character(),
  line = numeric(),
  stringsAsFactors = FALSE
)

for (geno_i in unique(soto.dat$Geno)) {
  data_i = soto.dat[soto.dat$Geno == geno_i, ]

  for (sex_j in c("F", "M")) {
    data_ij = data_i[data_i$sex == sex_j, ]
    if (nrow(data_ij) == 0) next
    if (length(unique(data_ij$Temp)) < 2) next

    data_means_ij = data_ij %>%
      group_by(Temp) %>%
      summarise(mean_tf = mean(time, na.rm = TRUE), .groups = "drop")

    if (nrow(data_means_ij) < 2) next

    lm_ij = lm(log10(mean_tf) ~ Temp, data = data_means_ij)
    a_hat = coef(lm_ij)[1]
    B_hat = coef(lm_ij)[2]

    real.pars.soto = rbind(
      real.pars.soto,
      data.frame(
        z = -1 / B_hat,
        sCTmax = -a_hat / B_hat,
        a = a_hat,
        B = B_hat,
        sex = sex_j,
        line = geno_i,
        stringsAsFactors = FALSE
      )
    )
  }
}
rownames(real.pars.soto) = NULL
```

```{r}
#refit PCA for these parameters
pca.males <- prcomp(real.pars.soto[real.pars.soto$sex=="M", c("a", "B")], scale. = TRUE)
pca.females <- prcomp(real.pars.soto[real.pars.soto$sex=="F", c("a", "B")], scale. = TRUE)

summary(pca.males)
summary(pca.females)

#redo PCA in B/sCTmax parameterization
pca.sCT.males <- prcomp(
  real.pars.soto[real.pars.soto$sex=="M", c("B", "sCTmax")],
  scale. = TRUE
)

pca.sCT.females <- prcomp(
  real.pars.soto[real.pars.soto$sex=="F", c("B", "sCTmax")],
  scale. = TRUE
)

#make phenotype files
DGRP2_PC1B = data.frame(DGRP=c(real.pars.soto$line[real.pars.soto$sex=="M"]),
                                Male=c(pca.males$x[,1]),
                                Female=c(pca.females$x[,1]))
DGRP2_PC2B = data.frame(DGRP=c(real.pars.soto$line[real.pars.soto$sex=="M"]),
                                Male=c(pca.males$x[,2]),
                                Female=c(pca.females$x[,2]))
DGRP2_zB = data.frame(DGRP=c(real.pars.soto$line[real.pars.soto$sex=="M"]),
                              Male=c(real.pars.soto$z[real.pars.soto$sex=="M"]),
                              Female=c(real.pars.soto$z[real.pars.soto$sex=="F"]))
DGRP2_sCTmaxB = data.frame(DGRP=c(real.pars.soto$line[real.pars.soto$sex=="M"]),
                                   Male=c(real.pars.soto$sCTmax[real.pars.soto$sex=="M"]),
                                   Female=c(real.pars.soto$sCTmax[real.pars.soto$sex=="F"]))
DGRP2_PC1_sCT = data.frame(
  DGRP = real.pars.soto$line[real.pars.soto$sex=="M"],
  Male = pca.sCT.males$x[,1],
  Female = pca.sCT.females$x[,1]
)

DGRP2_PC2_sCT = data.frame(
  DGRP = real.pars.soto$line[real.pars.soto$sex=="M"],
  Male = pca.sCT.males$x[,2],
  Female = pca.sCT.females$x[,2]
)
```

```{r}
aB_long <- data.frame(
  DGRP = real.pars.soto$line[real.pars.soto$sex=="M"],
  Male = real.pars.soto$a[real.pars.soto$sex=="M"],
  Female = real.pars.soto$a[real.pars.soto$sex=="F"]
) %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "sex",
               values_to = "aB")

BB_long <- data.frame(
  DGRP = real.pars.soto$line[real.pars.soto$sex=="M"],
  Male = real.pars.soto$B[real.pars.soto$sex=="M"],
  Female = real.pars.soto$B[real.pars.soto$sex=="F"]
) %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "sex",
               values_to = "BB")

PC1B_long <- DGRP2_PC1B %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "sex",
               values_to = "PC1B")

PC2B_long <- DGRP2_PC2B %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "sex",
               values_to = "PC2B")

zB_long <- DGRP2_zB %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "sex",
               values_to = "zB")

sCTmaxB_long <- DGRP2_sCTmaxB %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "sex",
               values_to = "sCTmaxB")

PC1C_long <- DGRP2_PC1_Centered %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "sex",
               values_to = "PC1_Centered")

zC_long <- DGRP2_z_Centered %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "sex",
               values_to = "z_Centered")

sCTmaxC_long <- DGRP2_sCTmax_Centered %>%
  pivot_longer(cols = c(Male, Female),
               names_to = "sex",
               values_to = "sCTmax_Centered")

PC1_sCT_long <- DGRP2_PC1_sCT %>%
  pivot_longer(
    cols = c(Male, Female),
    names_to = "sex",
    values_to = "PC1_sCT"
  )

PC2_sCT_long <- DGRP2_PC2_sCT %>%
  pivot_longer(
    cols = c(Male, Female),
    names_to = "sex",
    values_to = "PC2_sCT"
  )

final_df <- PC1B_long %>%
  left_join(PC2B_long, by = c("DGRP", "sex")) %>%
  left_join(zB_long, by = c("DGRP", "sex")) %>%
  left_join(sCTmaxB_long, by = c("DGRP", "sex")) %>%
  left_join(aB_long, by = c("DGRP", "sex")) %>%
  left_join(BB_long, by = c("DGRP", "sex")) %>%
  left_join(PC1C_long, by = c("DGRP", "sex")) %>%
  left_join(zC_long, by = c("DGRP", "sex")) %>%
  left_join(sCTmaxC_long, by = c("DGRP", "sex"))%>%
left_join(PC1_sCT_long, by=c("DGRP","sex")) %>%
left_join(PC2_sCT_long, by=c("DGRP","sex"))


for (i in 1:length(final_df$DGRP)){
  if (nchar(as.character(final_df$DGRP[i]))==2){
final_df$DGRP[i]=paste("DGRP_0",final_df$DGRP[i])}
  if (nchar(as.character(final_df$DGRP[i]))==1){
final_df$DGRP[i]=paste("DGRP_00",final_df$DGRP[i])}
  if (nchar(as.character(final_df$DGRP[i]))==3){
final_df$DGRP[i]=paste("DGRP_",final_df$DGRP[i])}
}

final_df$sex[final_df$sex=="Male"]="M"
final_df$sex[final_df$sex=="Female"]="F"

colnames(final_df) = c("DGRP","sex","PC1","PC2","z","sCTmax","a","B","PC1Centered","zCentered","sCTmaxCentered","PC1_sCTB","PC2_sCTB")


final_df$DGRP <- gsub("\\\\_", "_", final_df$DGRP)   # remove escape slashes
final_df$DGRP <- gsub(" ", "", final_df$DGRP)       # remove spaces


head(final_df)
```

```{r}
#this table is exported for DGRPool analysis
write.table(
  final_df,
  file = "DGRP2_all.txt",
  sep = "\t",
  row.names = FALSE,
  col.names = TRUE,
  quote = FALSE
)
```


```{r}
#Figure S5
pc_long <- final_df %>%
  dplyr::select(PC1, PC2, sex) %>%
  pivot_longer(cols = c(PC1, PC2),
               names_to = "PC",
               values_to = "value")

ggplot(pc_long, aes(x = value, fill = sex)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~PC, scales = "free") +
  theme_minimal() +
  labs(title = "", x = "Value", y = "Density")+
  xlim(c(-4,4))+
  theme(strip.text = element_text(size = 16, face = "bold"))
```
The following analysis uses the results of the DGRPool analysis

~~~~~~~~~~~~~~~~~~~~~~
Final GWAS output analysis
```{r}
# 1. Define traits and file paths
traits <- c("PC1", "PC2", "B", "a", "z", "sCTmax")

female_files <- c(
  "GWAS_Results/PC1_female.tsv",
  "GWAS_Results/PC2_female.tsv",
  "GWAS_Results/a_female.tsv",
  "GWAS_Results/B_female.tsv",
  "GWAS_Results/z_female.tsv",
  "GWAS_Results/sCTmax_female.tsv"
)

male_files <- c(
  "GWAS_Results/PC1_male.tsv",
  "GWAS_Results/PC2_male.tsv",
  "GWAS_Results/a_male.tsv",
  "GWAS_Results/B_male.tsv",
  "GWAS_Results/z_male.tsv",
  "GWAS_Results/sCTmax_male.tsv"
)

# Function to extract significant SNPs

get_sig_snps <- function(file, cutoff = 1e-5) {
  
  dt <- read.csv(file, sep = "\t", header = TRUE, stringsAsFactors = FALSE)
  
  unique(dt$ID[dt$P <= cutoff])
}

# Read all SNP sets
traits <- c("PC1", "PC2", "B", "a", "z", "sCTmax")

female_files_all <- list.files(pattern = "_female", full.names = TRUE)
male_files_all   <- list.files(pattern = "_male", full.names = TRUE)

# Match files to traits
female_files <- c(
  "GWAS_Results/PC1_female.tsv",
  "GWAS_Results/PC2_female.tsv",
  "GWAS_Results/B_female.tsv",
  "GWAS_Results/a_female.tsv",
  "GWAS_Results/z_female.tsv",
  "GWAS_Results/sCTmax_female.tsv"
)

male_files <- c(
  "GWAS_Results/PC1_male.tsv",
  "GWAS_Results/PC2_male.tsv",
  "GWAS_Results/B_male.tsv",
  "GWAS_Results/a_male.tsv",
  "GWAS_Results/z_male.tsv",
  "GWAS_Results/sCTmax_male.tsv"
)

names(female_files) <- traits
names(male_files)   <- traits

# Extract SNPs
female_snps <- lapply(female_files, get_sig_snps)
male_snps   <- lapply(male_files, get_sig_snps)

# Build matrix
n <- length(traits)
result_mat <- matrix("", n, n)
rownames(result_mat) <- traits
colnames(result_mat) <- traits

for (i in 1:n) {
  for (j in 1:n) {
    
    f_overlap <- length(intersect(female_snps[[i]], female_snps[[j]]))
    m_overlap <- length(intersect(male_snps[[i]], male_snps[[j]]))
    
    result_mat[i, j] <- paste0(f_overlap, "/", m_overlap)
  }
}

print(result_mat)
```

```{r}
# Females
female_extra <- setdiff(
  union(female_snps[["z"]], female_snps[["sCTmax"]]),
  female_snps[["PC1"]]
)
length(female_extra)

# Males
male_extra <- setdiff(
  union(male_snps[["z"]], male_snps[["sCTmax"]]),
  male_snps[["PC1"]]
)
length(male_extra)

# Females
female_extra <- setdiff(
  union(female_snps[["a"]], female_snps[["B"]]),
  female_snps[["PC1"]]
)
length(female_extra)

# Males
male_extra <- setdiff(
  union(male_snps[["a"]], male_snps[["B"]]),
  male_snps[["PC1"]]
)
length(male_extra)

# difference in total number of significant SNPS from a/B vs sCTmax/z parameterization
length(union(female_snps[["z"]], female_snps[["sCTmax"]]))-length(union(female_snps[["a"]], female_snps[["B"]]))
length(union(male_snps[["z"]], male_snps[["sCTmax"]]))-length(union(male_snps[["a"]], male_snps[["B"]]))

# of significant SNPS associated with z, but not sCTmax
length(union(female_snps[["z"]],female_snps[["sCTmax"]]))
length(union(female_snps[["a"]],female_snps[["B"]]))
```

Same thing but for Table S1
```{r}

# Traits to compare
traits <- c("PC1", "PC2", "B", "sCTmax")

# GWAS result files
female_files <- c(
  "PC1_female_sCT.tsv",
  "PC2_female_sCT.tsv",
  "B_female.tsv",
  "sCTmax_female.tsv"
)

male_files <- c(
  "PC1_male_sCT.tsv",
  "PC2_male_sCT.tsv",
  "B_male.tsv",
  "sCTmax_male.tsv"
)

names(female_files) <- traits
names(male_files) <- traits

# Function to get significant SNPs
get_sig_snps <- function(file, cutoff = 1e-5) {

  dt <- read.csv(
    file,
    sep = "\t",
    header = TRUE,
    stringsAsFactors = FALSE
  )

  unique(dt$ID[dt$P <= cutoff])
}

# Read SNP sets
female_snps <- lapply(female_files, get_sig_snps)
male_snps   <- lapply(male_files, get_sig_snps)

# Build overlap matrix
n <- length(traits)

result_mat <- matrix("", nrow = n, ncol = n)
rownames(result_mat) <- traits
colnames(result_mat) <- traits

for (i in seq_len(n)) {
  for (j in seq_len(n)) {

    f_overlap <- length(
      intersect(female_snps[[i]],
                female_snps[[j]])
    )

    m_overlap <- length(
      intersect(male_snps[[i]],
                male_snps[[j]])
    )

    result_mat[i, j] <- paste0(f_overlap, "/", m_overlap)
  }
}

print(result_mat)


```

```{r}
#Tests for whether # of significant SNPs deviates from what is expected under random chance
gwas_null_p <- function(Nsig, Ntests=2000000, alpha=1e-5){
  pbinom(
    Nsig-1,
    size=Ntests,
    prob=alpha,
    lower.tail=FALSE
  )
}
#Females
#PC1; PC2; B; a
gwas_null_p(79); gwas_null_p(14); gwas_null_p(98); gwas_null_p(75)
#sCTmax; PC1; PC2 (sCTmax/B)
gwas_null_p(122); gwas_null_p(132); gwas_null_p(15)
#Males
#PC1; PC2; B; a
gwas_null_p(28); gwas_null_p(12); gwas_null_p(27); gwas_null_p(27)
#sCTmax; PC1; PC2 (sCTmax/B)
gwas_null_p(122); gwas_null_p(28); gwas_null_p(28)
```


~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~


Figure S2
```{r}
rm(list = ls())
library(tidyverse)
```


```{r}
soto.dat = read.csv("Soto_etal_data/KO_4temp_100DGRP.csv")


#format the data for the model
soto.dat$time=soto.dat$KO/60
soto.dat$log_tf=log10(soto.dat$time)
soto.dat$Temp = soto.dat$temp
soto.dat$line=soto.dat$Geno
```

```{r}
#first, we center with respect to x to reduce unidentifiability in a/B parameterization (centering around a)
Tbar = mean(soto.dat$Temp, na.rm = TRUE)
soto.dat.centered <- soto.dat %>%
  mutate(
    Temp_raw = Temp,
    Temp = Temp_raw - Tbar
  )

centered.pars.soto <- data.frame(
  z = numeric(),
  sCTmax = numeric(),
  a = numeric(),
  B = numeric(),
  sex = character(),
  line=numeric(),
  stringsAsFactors = FALSE
)

for (geno_i in unique(soto.dat.centered$Geno)) {
  data_i <- soto.dat.centered[soto.dat.centered$Geno == geno_i, ]
  for (sex_j in c("F", "M")) {
    data_ij <- data_i[data_i$sex == sex_j, ]
    data_means_ij = data_ij %>%
  group_by(Temp) %>%
  summarise(
    mean_tf = mean(time, na.rm = TRUE),
    .groups = "drop"
  )
    lm_ij <- lm(log10(mean_tf) ~ Temp, data = data_means_ij)
    a_hat <- coef(lm_ij)[1]
    B_hat <- coef(lm_ij)[2]
    sCTmax_ij <- -a_hat / B_hat
    #version of sCTmax with NEW PARAMETERIZATION
    z_ij <- -1 / B_hat
    centered.pars.soto <- rbind(
      centered.pars.soto,
      data.frame(z = z_ij, sCTmax = sCTmax_ij,a=a_hat, B=B_hat, sex = sex_j,line=geno_i, stringsAsFactors = FALSE))
  }}
```

```{r}
library(dplyr)
library(ggplot2)

# correlation for each sex
cor.dat <- centered.pars.soto %>%
  group_by(sex) %>%
  summarise(
    rho = cor(a, B, use = "complete.obs"),
    x = min(a) + 0.05 * diff(range(a)),
    y = max(B) - 0.05 * diff(range(B)),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("\u03C1 = ", round(rho, 2))
  )

ggplot(centered.pars.soto, aes(x = a, y = B)) +
  stat_density_2d_filled() +
  geom_text(
    data = cor.dat,
    aes(x = 1.75, y = -0.12, label = label),
    inherit.aes = FALSE,
    color = "white",
    size = 14 / .pt   # 14 pt text
  ) +
  facet_wrap(
    ~sex,
    nrow = 1,
    labeller = labeller(sex = c(F = "Females",
                                M = "Males"))
  ) +
  theme_minimal() +
  labs(
    x = "Intercept (\u03b1)",
    y = "Slope (\u03b2)"
  ) +
  theme(
    legend.position = "none",
    axis.title = element_text(size = 14),
    strip.text = element_text(size = 14, face = "bold")
  )
```

```{r}
#extract centered parameter distribution stats for power analysis
mean(centered.pars.soto$a);var(centered.pars.soto$a)
mean(centered.pars.soto$B);var(centered.pars.soto$B)
```

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Figure S3

```{r}
rm(list = ls())
source("identifiability_functions.R")
library(patchwork)
library(tidyverse)

soto.dat = read.csv("Soto_etal_data/KO_4temp_100DGRP.csv")
head(soto.dat)
#format the data for the model
soto.dat$time=soto.dat$KO/60
soto.dat$log_tf=log10(soto.dat$time)
soto.dat$Temp = soto.dat$temp
soto.dat$line=soto.dat$Geno
soto.dat.m=soto.dat[soto.dat$sex=="M",]
soto.dat.f=soto.dat[soto.dat$sex=="F",]
```

```{r}
  soto_result.m.cor <- reaction_norm_null_test(
    data = soto.dat.m,
    response = "log_tf",
    predictor = "Temp",
    group = "line",
    statistic = c("correlation"),
    alternative="two.sided",
    n_sims = 5000,
    center_predictor = FALSE,  # preserves intercept at Temp == 0
    seed = 1
  )
  soto_result.m.cov <- reaction_norm_null_test(
    data = soto.dat.m,
    response = "log_tf",
    predictor = "Temp",
    group = "line",
    statistic = c("covariance"),
    alternative="two.sided",
    n_sims = 5000,
    center_predictor = FALSE,  # preserves intercept at Temp == 0
    seed = 1
  )

  print(soto_result.m.cor)
  print(soto_result.m.cov)
  wrap_plots((plot(soto_result.m.cor)+xlab("Correlation")),(plot(soto_result.m.cov)+xlab("Covariance")))+
                plot_layout(guides = "collect") &
                theme(legend.position = "bottom")&
               theme(strip.text = element_blank(), strip.background = element_blank())
  
  #females
    soto_result.f.cor <- reaction_norm_null_test(
    data = soto.dat.f,
    response = "log_tf",
    predictor = "Temp",
    group = "line",
    statistic = "correlation",
    alternative="two.sided",
    n_sims = 5000,
    center_predictor = FALSE,  # preserves intercept at Temp == 0
    seed = 1
  )
    
        soto_result.f.cov <- reaction_norm_null_test(
    data = soto.dat.f,
    response = "log_tf",
    predictor = "Temp",
    group = "line",
    statistic = "covariance",
    alternative="two.sided",
    n_sims = 5000,
    center_predictor = FALSE,  # preserves intercept at Temp == 0
    seed = 1
  )

   print(soto_result.f.cor)
  print(soto_result.f.cov)
  wrap_plots((plot(soto_result.f.cor)+xlim(c(-1,-0.98))),
             (plot(soto_result.f.cov)+ylab("")),
             (plot(soto_result.m.cor)+xlab("Correlation")+xlim(c(-1,-0.98))),(plot(soto_result.m.cov)+xlab("Covariance")+ylab("")),nrow=2)+
                plot_annotation(tag_levels = "A")+
                plot_layout(guides = "collect") &
                theme(legend.position = "bottom")&
               theme(strip.text = element_blank(), strip.background = element_blank())&ylim(c(0,2000))

```

```{r}
#covariance
  soto_result.m <- reaction_norm_null_test(
    data = soto.dat.m,
    response = "log_tf",
    predictor = "Temp",
    group = "line",
    statistic = "covariance",
    alternative="two.sided",
    n_sims = 5000,
    center_predictor = FALSE,  # preserves intercept at Temp == 0
    seed = 1
  )

  print(soto_result.m)
  
    soto_result.f <- reaction_norm_null_test(
    data = soto.dat.f,
    response = "log_tf",
    predictor = "Temp",
    group = "line",
    statistic = "covariance",
    alternative="two.sided",
    n_sims = 5000,
    center_predictor = FALSE,  # preserves intercept at Temp == 0
    seed = 1
  )

  print(soto_result.f)
```

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

Figure S8

```{r}
library(mgcv)
```

```{r}
rm(list = ls())
#N=number of lines
N=100
#Panel 1
set.seed(11)
#Here we are testing the effect of n (samples per temp) on model identifiability
#We will choose 5 equally spaced temps to take data at. Using a delta T with optimal performance in fig 1: 
N_T=5
delta.T=8
temps = seq(20,44,0.1)
sCTmax.mean = 46
sCTmax.sd = 1
z.mean = 4.5
z.sd = 1
#correlation between slope and intercept for between line variation (r.genetic) and between individual variation (r.plastic)
#making it 0 to prove a point, but you could move it around.
r.genetic = 0
r.plastic=r.genetic
#Constant variance across temperatures isn't a realistic assumption (Bullard et al, 2026) but it still demonstrates our point
#It also is a simplifcation to assume non-stochastic errors, but again, that's ok for these purposes
var.mean = 0.01
sd.mean=sqrt(var.mean)
cov.sctmax.z = r.genetic*z.sd*sCTmax.sd
cov.indiv = r.plastic*z.sd*sCTmax.sd
mu=c(sCTmax.mean,z.mean)
sigma = matrix(data=c(sCTmax.sd^2,cov.sctmax.z,cov.sctmax.z,z.sd^2),nrow=2)
sigma.indiv = matrix(data=c(sCTmax.sd^2,cov.indiv,cov.indiv,z.sd^2),nrow=2)
temp_midpoint = (temps[1]+temps[length(temps)])/2
temp.list.p1 = round(
  seq(temp_midpoint + delta.T/2,
      temp_midpoint - delta.T/2,
      length.out = N_T), 1)
n.list = c(1:100)

#use this for distribution comparison throughout
kl_mvn <- function(mu_hat, Sigma_hat, mu, Sigma) {
  k <- length(mu)
  0.5 * (log(det(Sigma) / det(Sigma_hat)) -k +sum(diag(solve(Sigma) %*% Sigma_hat)) +t(mu - mu_hat) %*% solve(Sigma) %*% (mu - mu_hat))
}

variable.n.df = data.frame(kl=rep(NA,length(n.list)),n=n.list)

for (k in 1:length(n.list)){
  n_k=n.list[k]
  #Here we will generate the data
TDT.df_k = data.frame(log10_tf=c(), temp=c(), line=c(),z=c(),sCTmax=c())
  for (i in 1:N){
    #genetic effects on TDT parameters (constant for a given line)
    #we are just going to assume that 50% of the variation in TDT parameters is explained by genetics and 50% is individual (plastic), which is incorporated in the TDT.df_ij line. These are generalizations that won't affect our qualitative results
    sCTmax_i = rnorm(1,sCTmax.mean,sCTmax.sd/2)
    z_i = rnorm(1,z.mean,z.sd/2)
    for (j in 1:length(temp.list.p1)){
      temp_j=temp.list.p1[j]
      #here we incorporate individual environmental (plastic) variation into our TDT parameters
        sCTmax_ij = rnorm(n_k,mean=sCTmax_i,sCTmax.sd/2)
        z_ij = rnorm(n_k,z_i,z.sd)
      TDT.df_ij = data.frame(
        #here we simulate the TDT data for a given line at a given temp. Note that individual variation in TDT parameters is incorporated here, as well as a general error term
        log10_tf = rnorm(rep(1,n_k),(sCTmax_ij-temp_j)/z_ij,rep(sd.mean,n_k)),
                             temp=rep(temp_j,n_k),line=rep(i,n_k))
      TDT.df_k = rbind(TDT.df_k,TDT.df_ij)
    }
  }
  #analyze the data for this n
  
  #run an lm to estimate the original pars
  lines.df_k = data.frame(line=c(), slope = c(), intercept = c())
  #fit least squares regression for each line
  for (j in 1:N){
    data_j = TDT.df_k[TDT.df_k$line==j,]
    lm_j = lm(log10_tf~temp,data=data_j)
    lines.df_k = rbind(lines.df_k, data.frame(line=j,slope=coefficients(lm_j)[2],intercept=coefficients(lm_j)[1]))
  }
  #reparameterize
  lines.df_k$sCTmax = -lines.df_k$intercept/lines.df_k$slope
  lines.df_k$z = -1/lines.df_k$slope
  #KL divergence
  mu_hat = colMeans(lines.df_k[,c("sCTmax","z")])
  print(colMeans(lines.df_k[,c("slope","intercept")]))
  sigma_hat = cov(lines.df_k[,c("sCTmax","z")])
variable.n.df$kl[k] = kl_mvn(mu_hat,sigma_hat,mu,sigma)
}
```

```{r}
#panel 1:
p1mod=lm(log(kl)~n,data=variable.n.df)
pvalue=summary(p1mod)$coefficients[2,4]

p1=ggplot(data=variable.n.df)+
  geom_point(aes(y=log(kl),x=log10(n)))+
  theme_classic()+
  geom_smooth(aes(y=log(kl),x=log10(n)),method="lm",se=FALSE)+
  xlab(expression(log[10] * "(" * n[per~temp] * ")"))+
  ylab("ln(KL)")+
  xlab(expression(n["per temp"])) +
  annotate("text", x = 1, y = 7.5,
           label = paste0("p = ", round(pvalue, 2)),
           hjust = 1.1, vjust = 1.5,size=5)

```

```{r}
#panel 2
#here we are looking for the effect of number of temps
#I chose 180 because it is divisible by all of the N_T integers below, and it is on par with the number of samples researchers typically use for a TDT study
#n*N_T
n.total=180
N_T.list = c(2,3,4,5,6,9,10,12)
delta.T = 8
#This analysis is inherently limited because we can't sample at a fine resolution of N_T because it must be an integer and n must also be an integer. So, to reduce sampling error, we can either run this for many seeds and average our results, or we can increase N (number of lines). I'll opt for the former to maintain consistency with previous trials. 

#number of seeds to try
n_s=200

variable.N_T.df=data.frame(kl=c(),N_T=c())

#I know 4 for loops is a lot (don't tell the computer scientists), but it runs fairly quickly, and I feel that it shows the process more clearly than other methods

for (s in 1:n_s){
set.seed(s)
  variable.N_T.df_s = data.frame(kl=rep(NA,length(N_T.list)),N_T=N_T.list)
  
for (k in 1: length(N_T.list)){
  N_T_k=N_T.list[k]
  temp.list_k = seq(temp_midpoint-delta.T/2,temp_midpoint+delta.T/2,length.out=N_T_k)
  n_k=n.total/N_T_k
  TDT.df_k = data.frame(log10_tf=c(), temp=c(), line=c(),z=c(),sCTmax=c())
  
    for (i in 1:N){
      sCTmax_i = rnorm(1,sCTmax.mean,sCTmax.sd/2)
      z_i = rnorm(1,z.mean,z.sd/2)
      
      for (j in 1:length(temp.list_k)){
        temp_j=temp.list_k[j]
        #here we incorporate individual environmental (plastic) variation into our TDT parameters
          sCTmax_ij = rnorm(n_k,mean=sCTmax_i,sCTmax.sd/2)
          z_ij = rnorm(n_k,z_i,z.sd)
        TDT.df_ij = data.frame(
          #here we simulate the TDT data for a given line at a given temp. Note that individual variation in TDT parameters is incorporated here, as well as a general error term
          log10_tf = rnorm(rep(1,n_k),(sCTmax_ij-temp_j)/z_ij,rep(sd.mean,n_k)),
                               temp=rep(temp_j,n_k),line=rep(i,n_k))
        TDT.df_k = rbind(TDT.df_k,TDT.df_ij)
    }
    }
  #analyze for this N_T
  
  #run an lm to estimate the original pars
  lines.df_k = data.frame(line=c(), slope = c(), intercept = c())
  #fit least squares regression for each line
  for (j in 1:N){
    data_j = TDT.df_k[TDT.df_k$line==j,]
    lm_j = lm(log10_tf~temp,data=data_j)
    lines.df_k = rbind(lines.df_k, data.frame(line=j,slope=coefficients(lm_j)[2],intercept=coefficients(lm_j)[1]))
  }
  #reparameterize
  lines.df_k$sCTmax = -lines.df_k$intercept/lines.df_k$slope
  lines.df_k$z = -1/lines.df_k$slope
  #KL divergence
  mu_hat = colMeans(lines.df_k[,c("sCTmax","z")])
  sigma_hat = cov(lines.df_k[,c("sCTmax","z")])
variable.N_T.df_s$kl[k] = kl_mvn(mu_hat,sigma_hat,mu,sigma)
}
  variable.N_T.df = rbind(variable.N_T.df,variable.N_T.df_s)
  }

```

```{r}
#panel 2:

gammod = gam(log(kl) ~ s(N_T, k = 3), data = variable.N_T.df)
pvalue = summary(gammod)$s.table[1, "p-value"]



p2=ggplot(variable.N_T.df) +
  geom_boxplot(
    aes(x = N_T, y = log(kl), group = N_T),
    width = 0.8) +
  scale_x_continuous(breaks = sort(unique(variable.N_T.df$N_T))) +
  theme_classic()+
geom_smooth(aes(x = N_T, y = log(kl),group=1),method = "gam",
    formula = y ~ s(x, k =3),se=FALSE)+
  ylab("ln(KL)")+
  xlab(expression(n[temps]))+
  annotate("text", x = 7, y = 10,
           label = paste0("p = ", round(pvalue, 2)),
           hjust = 1.1, vjust = 1.5, size=5)



#Apparently, it is best to put all of your eggs in *two* baskets: estimations are more accurate when we sample more points from fewer, farther apart temperatures. Intuitively this makes sense because it gives each point maximum leverage. Additionally, having more samples at your temperatures of interest helps you more accurately depict the shape of the distribution, if that is something you're interested in. 
```

```{r}
#panel 3
#Here we want to test temperature spacing. If we use more than two temperatures, should they be clustered at high or low temps?
#We will use 3 temps: 2 fixed 8C apart, and the third will vary between the two
delta.T=12
Tmin=temp_midpoint-delta.T/2
Tmax=temp_midpoint+delta.T/2
temp3.list = seq(Tmin,Tmax,0.01)

variable.temp3.df = data.frame(kl=rep(NA,length(temp3.list)),temp3=temp3.list)

for (k in 1: length(temp3.list)){
  temp3_k = temp3.list[k]
  temp.list_k = c(Tmin,temp3_k,Tmax)
  
  TDT.df_k = data.frame(log10_tf=c(), temp=c(), line=c(),z=c(),sCTmax=c())
  
    for (i in 1:N){
      #line (genetic) variation
      sCTmax_i = rnorm(1,sCTmax.mean,sCTmax.sd/2)
      z_i = rnorm(1,z.mean,z.sd/2)
      
      for (j in 1:length(temp.list_k)){
        temp_j=temp.list_k[j]
        #here we incorporate individual environmental (plastic) variation into our TDT parameters
          sCTmax_ij = rnorm(n_k,mean=sCTmax_i,sCTmax.sd/2)
          z_ij = rnorm(n_k,z_i,z.sd)
        TDT.df_ij = data.frame(
          #here we simulate the TDT data for a given line at a given temp. Note that individual variation in TDT parameters is incorporated here, as well as a general error term
          log10_tf = rnorm(rep(1,n_k),(sCTmax_ij-temp_j)/z_ij,rep(sd.mean,n_k)),
                               temp=rep(temp_j,n_k),line=rep(i,n_k))
        TDT.df_k = rbind(TDT.df_k,TDT.df_ij)
    }}
  
  #analyze for this temp3
  
  #run an lm to estimate the original pars
  lines.df_k = data.frame(line=c(), slope = c(), intercept = c())
  #fit least squares regression for each line
  for (j in 1:N){
    data_j = TDT.df_k[TDT.df_k$line==j,]
    lm_j = lm(log10_tf~temp,data=data_j)
    lines.df_k = rbind(lines.df_k, data.frame(line=j,slope=coefficients(lm_j)[2],intercept=coefficients(lm_j)[1]))
  }
  #reparameterize
  lines.df_k$sCTmax = -lines.df_k$intercept/lines.df_k$slope
  lines.df_k$z = -1/lines.df_k$slope
  #KL divergence
  mu_hat = colMeans(lines.df_k[,c("sCTmax","z")])
  sigma_hat = cov(lines.df_k[,c("sCTmax","z")])
variable.temp3.df$kl[k] = kl_mvn(mu_hat,sigma_hat,mu,sigma)
  
}

```


```{r}
#panel 3
#Here we test whether the vertex of this parabola is significantly different from the midpoint of these two temperatures. This would indicate that one side is preferred over the other.
# fit raw quadratic
# Fit raw quadratic 
fit <- lm(log(kl) ~ temp3 + I(temp3^2), data = variable.temp3.df)
beta <- coef(fit)
beta1 <- beta["temp3"]
beta2 <- beta["I(temp3^2)"]

# Vertex and SE via delta method
vcov_mat <- vcov(fit)
dx_db1 <- -1 / (2*beta2)
dx_db2 <- beta1 / (2*beta2^2)
var_vertex <- dx_db1^2 * vcov_mat["temp3", "temp3"] +
              dx_db2^2 * vcov_mat["I(temp3^2)", "I(temp3^2)"] +
              2*dx_db1*dx_db2*vcov_mat["temp3", "I(temp3^2)"]
se_vertex <- sqrt(var_vertex)
x_vertex <- -beta1/(2*beta2)
ci_lower <- x_vertex - 1.96*se_vertex
ci_upper <- x_vertex + 1.96*se_vertex

# Prepare fitted line
x_vals <- seq(min(variable.temp3.df$temp3), max(variable.temp3.df$temp3), length.out = 100)
y_vals <- beta[1] + beta1 * x_vals + beta2 * x_vals^2
fit_df <- data.frame(x = x_vals, y = y_vals)

z_value <- (x_vertex - temp_midpoint) / se_vertex
p_value <- 2 * pnorm(-abs(z_value))
p_value

# Plot
p3=ggplot(variable.temp3.df, aes(x = temp3, y = log(kl))) +
  geom_point(alpha = 0.3) +
  geom_line(data = fit_df, aes(x = x, y = y), color = "blue", linewidth = 1) +
  geom_vline(xintercept = x_vertex, color = "purple", linetype = "dashed") +
  geom_rect(aes(xmin = ci_lower, xmax = ci_upper, ymin = -Inf, ymax = Inf),
            fill = "purple", alpha = 0.005) +
  labs(
    x = expression("Temperature (" * degree * "C)"),
    y = "ln(KL)") +
  theme_classic()+
  annotate("text", x = 35, y = 7.5, 
           label = paste("italic(p) == ", round(p_value,2)), 
           parse = TRUE, size = 5)
middle.temp.mod = lm(log(kl)~sCTmax.diff,data=variable.middle.temp.df)
#not significantly different: It doesn't really matter which side the extra temperature is on
```

```{r}
#panel 4
#Here we are testing whether shifting all of the temperatures left or right matters
delta.T=6 #using a smaller delta T so we have more room to work with
N_T=3
middle.temp.p4 = seq(30,40,0.01)

variable.middle.temp.df = data.frame(kl=rep(NA,length(middle.temp.p4)),middle.temp=middle.temp.p4)

for (k in 1:length(middle.temp.p4)){
  middle.temp.p4_k = middle.temp.p4[k]
  temps.list_k = c(middle.temp.p4_k-delta.T/2,middle.temp.p4_k,middle.temp.p4_k+delta.T/2)
  TDT.df_k = data.frame(log10_tf=c(), temp=c(), line=c(),z=c(),sCTmax=c())
  
  for (i in 1:N){
      #line (genetic) variation
      sCTmax_i = rnorm(1,sCTmax.mean,sCTmax.sd/2)
      z_i = rnorm(1,z.mean,z.sd/2)
      
      for (j in 1:length(temp.list_k)){
        temp_j=temp.list_k[j]
        #here we incorporate individual environmental (plastic) variation into our TDT parameters
          sCTmax_ij = rnorm(n_k,mean=sCTmax_i,sCTmax.sd/2)
          z_ij = rnorm(n_k,z_i,z.sd)
        TDT.df_ij = data.frame(
          #here we simulate the TDT data for a given line at a given temp. Note that individual variation in TDT parameters is incorporated here, as well as a general error term
          log10_tf = rnorm(rep(1,n_k),(sCTmax_ij-temp_j)/z_ij,rep(sd.mean,n_k)),
                               temp=rep(temp_j,n_k),line=rep(i,n_k))
        TDT.df_k = rbind(TDT.df_k,TDT.df_ij)
    }}
  
  #analyze for this middle.temp
  
  #run an lm to estimate the original pars
  lines.df_k = data.frame(line=c(), slope = c(), intercept = c())
  #fit least squares regression for each line
  for (j in 1:N){
    data_j = TDT.df_k[TDT.df_k$line==j,]
    lm_j = lm(log10_tf~temp,data=data_j)
    lines.df_k = rbind(lines.df_k, data.frame(line=j,slope=coefficients(lm_j)[2],intercept=coefficients(lm_j)[1]))
  }
  #reparameterize
  lines.df_k$sCTmax = -lines.df_k$intercept/lines.df_k$slope
  lines.df_k$z = -1/lines.df_k$slope
  #KL divergence
  mu_hat = colMeans(lines.df_k[,c("sCTmax","z")])
  sigma_hat = cov(lines.df_k[,c("sCTmax","z")])
variable.middle.temp.df$kl[k] = kl_mvn(mu_hat,sigma_hat,mu,sigma)
  
  
}
```

```{r}
#panel 4:
p_value=summary(middle.temp.mod)$coefficients[2,4]
p_value
variable.middle.temp.df$max.temp = variable.middle.temp.df$middle.temp+delta.T/2
variable.middle.temp.df$sCTmax.diff = sCTmax.mean-variable.middle.temp.df$max.temp
p4=ggplot(data=variable.middle.temp.df)+
  geom_point(aes(y=log(kl),x=sCTmax.diff))+
  geom_smooth(aes(y=log(kl),x=sCTmax.diff),method="lm",se=FALSE)+
  theme_classic()+
xlab(expression("Left temperature shift (" * degree * "C)"))+
  annotate("text", x = 10, y = 7.5, 
           label = paste("italic(p) == ", round(p_value,2)), 
           parse = TRUE, size = 5)+
  ylab("ln(KL)")
middle.temp.mod = lm(log(kl)~sCTmax.diff,data=variable.middle.temp.df)

#no difference. You don't need to avoid high/low temps. Although, this is something that might require consideration of practical constraints, and nonconformities to our simulation structure, as outliers will become larger as you near t=0 (skew may increase, model effectiveness decrease)
```

```{r}
#Full figure
wrap_plots(p1,p2,p3,p4,nrow=2)+
    plot_annotation(tag_levels = "A")&
  theme(
    axis.title = element_text(size = 12),
    axis.text  = element_text(size = 10),  # Keeps tick text above 6pt minimum
    plot.margin = margin(5, 5, 5, 5, "mm"))

```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Figure S6
Let's check the studies that Gunderson found to be artificial
```{r}
rm(list = ls())
source("identifiability_functions.R")
library(patchwork)
library(tidyverse)
#check which ones have more than one species
file <- "compiled_reanalysis_data_V2.xlsx"

read_sheet_clean <- function(sheet) {
  df <- readxl::read_excel(file, sheet = sheet)
  names(df) <- make.names(names(df))
  df
}

datasets_init <- list(
  vanHeerwaarden = read_sheet_clean("van Heerwaarden 2016"),
  Deery = read_sheet_clean("Deery 2021"),
  Phillips = read_sheet_clean("Phillips 2016"),
  Gilbert = read_sheet_clean("Gilbert 2019"),
  Sasaki19 = read_sheet_clean("Sasaki 2019"),
  Sasaki20 = read_sheet_clean("Sasaki 2020"),
  #only SP pop control, PG pop, control
  Sasaki21 = read_sheet_clean("Sasaki 2021"),
  Noh = read_sheet_clean("Noh 2017"),
  Oneillcco = read_sheet_clean("Oneil 2021 cco"),
  Oneillccrt = read_sheet_clean("Oniel 2021 ccrt"),
  Nguyen = read_sheet_clean("Nguyen 2019"),
  Mottola = read_sheet_clean("Mottola 2022"),
  Morgan = read_sheet_clean("Morgan 2018")
)

datasets = list( 
  Mottola = datasets_init$Mottola,
                  Morgan.M = datasets_init$Morgan[datasets_init$Morgan$sex=="male",],
                  Morgan.F = datasets_init$Morgan[datasets_init$Morgan$sex=="female",],
                  Phillips_O1 = datasets_init$Phillips[datasets_init$Phillips$observer==1,],
  Phillips_O2 = datasets_init$Phillips[datasets_init$Phillips$observer==2,],
  Gilbert = datasets_init$Gilbert,
                  vanHeerwaarden = datasets_init$vanHeerwaarden,
                  SasakiCT_control = datasets_init$Sasaki21[(datasets_init$Sasaki21$pop=="CT" & datasets_init$Sasaki21$treat=="control"),],
                  SasakiCT_warming = datasets_init$Sasaki21[(datasets_init$Sasaki21$pop=="CT" & datasets_init$Sasaki21$treat=="warming"),],
                  SasakiSP_control = datasets_init$Sasaki21[(datasets_init$Sasaki21$pop=="SP" & datasets_init$Sasaki21$treat=="control"),],
                  SasakiSP_warming = datasets_init$Sasaki21[(datasets_init$Sasaki21$pop=="SP" & datasets_init$Sasaki21$treat=="warming"),],
                  SasakiPG_control = datasets_init$Sasaki21[(datasets_init$Sasaki21$pop=="CT" & datasets_init$Sasaki21$treat=="control"),],
                  SasakiPG_warming = datasets_init$Sasaki21[(datasets_init$Sasaki21$pop=="PG" & datasets_init$Sasaki21$treat=="warming"),],
                  Deery_caroline = datasets_init$Deery[datasets_init$Deery$species=="carolinensis",],
                  Deery_sagrei = datasets_init$Deery[datasets_init$Deery$species=="sagrei",],
                  Sasaki19 = datasets_init$Sasaki19,
                  Sasaki20 = datasets_init$Sasaki20,
                  Noh_low = datasets_init$Noh[datasets_init$Noh$treatment=="low",],
                  Noh_mid = datasets_init$Noh[datasets_init$Noh$treatment=="mid",],
                  Noh_high = datasets_init$Noh[datasets_init$Noh$treatment=="high",],
                  Oneillcco15 = datasets_init$Oneillcco[datasets_init$Oneillcco$treatment=="15C",],
                  Oneillcco25 = datasets_init$Oneillcco[datasets_init$Oneillcco$treatment=="25C",],
                  Oneillccrt15 = datasets_init$Oneillccrt[datasets_init$Oneillccrt$treatment=="15C",],
                  Oneillccrt25 = datasets_init$Oneillccrt[datasets_init$Oneillccrt$treatment=="25C",],
                  Nguyen = datasets_init$Nguyen
                  )
#remove unaltered columns from Nguyen so they aren't picked up by the cleaner function
datasets$Nguyen=datasets$Nguyen[,(colnames(datasets$Nguyen)%in%c("baseline.ccrt.", "acclimated.ccrt.")==FALSE)]

prep_data <- function(df, name) {

  possible_pairs <- list(
    c("ctmax1", "ctmax2"),
    c("t18", "t22"),
    c("ct1", "ct2"),
    c("baseline.ccrt.", "acclimated.ccrt."),
    c("cor.cco1","cor.cco2"),
    c("ccrt1","ccrt2"),
    c("cor.acclimated","cor.baseline"),
    c("CTmax1","CTmax2"),
    c("w1","w2"),
    c("basal", "hard")
  )

  matching_pairs <- Filter(
    function(pair) all(pair %in% names(df)),
    possible_pairs
  )

  if (length(matching_pairs) == 0L) {
    stop("Could not detect paired columns in ", name, ".")
  }

  if (length(matching_pairs) > 1L) {
    warning(
      "More than one measurement pair was found in ",
      name,
      ". Using ",
      paste(matching_pairs[[1L]], collapse = " and "),
      "."
    )
  }

  pair <- matching_pairs[[1L]]

  df |>
    dplyr::transmute(
      ID = factor(seq_len(dplyr::n())),
      y1 = as.numeric(.data[[pair[1L]]]),
      y2 = as.numeric(.data[[pair[2L]]])
    ) |>
    tidyr::pivot_longer(
      cols = c(y1, y2),
      names_to = "occasion",
      values_to = "trait"
    ) |>
    dplyr::mutate(
      stage = ifelse(occasion == "y1", 0, 1)
    ) |>
    dplyr::select(ID, stage, trait) |>
    dplyr::filter(
      is.finite(stage),
      is.finite(trait)
    ) |>
    dplyr::group_by(ID) |>
    dplyr::filter(
      dplyr::n() == 2L,
      dplyr::n_distinct(stage) == 2L
    ) |>
    dplyr::ungroup() |>
    droplevels()
}

long_datasets <- purrr::imap(datasets, prep_data)

design_summary <- purrr::imap_dfr(
  long_datasets,
  function(df, dataset_name) {
    tibble(
      dataset = dataset_name,
      n_ID = dplyr::n_distinct(df$ID),
      n_observations = nrow(df),
      observations_per_ID = nrow(df) / dplyr::n_distinct(df$ID)
    )
  }
)
```


```{r}
settings <- list(
  statistic = "correlation",
  n_sims = 1000,
  center_predictor = TRUE,
  test_uncentered_intercept = TRUE,
  seed = 1,
  min_group_n = 2,
  alternative = "two.sided"
)

results <- imap(long_datasets, \(d, nm) {

  do.call(
    reaction_norm_null_test,
    c(
      list(
        data = d,
        response = "trait",
        predictor = "stage",
        group = "ID"
      ),
      settings
    )
  )
})

ncol <- 5

plots <- imap(
  results,
  function(res, nm) {

    i <- which(names(results) == nm)

    left_col  <- ((i - 1) %% ncol) == 0
    bottom_row <- i > (length(results) - ncol)

    p <- plot(res) +
      ggtitle(nm) +
      coord_cartesian(
        xlim = c(-1, 1),
        ylim = c(0, 200)
      ) +
      annotate(
        "text",
        x = Inf,
        y = Inf,
        label = paste0(
          "p = ",
          round(res$summary$p_value[1], 4)
        ),
        hjust = 1.1,
        vjust = 1.5
      )

    if (!left_col) {
      p <- p +
        theme(
          axis.title.y = element_blank(),
          axis.text.y  = element_blank(),
          axis.ticks.y = element_blank()
        )
    }

if (bottom_row) {
  p <- p +
    labs(x = "Correlation")   # your desired x-axis title
} else {
  p <- p +
    labs(x = NULL) +
    theme(
      axis.title.x = element_blank(),
      axis.text.x  = element_blank(),
      axis.ticks.x = element_blank()
    )
}
#remove facet strips (axis label)
    p = p+theme(strip.text = element_blank(),
strip.background = element_blank())
    p
  }
)



wrap_plots(plots,ncol=5)+
  plot_annotation(tag_levels = "A")+
plot_layout(guides = "collect") &
theme(legend.position = "bottom")
```
~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Figure S7
```{r}
rm(list = ls())
#Great Tits analysis
library(lme4)
library(ggplot2)
library(dplyr)
library(patchwork)
library(MASS)

source("identifiability_functions.R")

dat <- read.csv("Great_Tits.csv")

dat$stage <- dat$NestStage2       # centered predictor 
dat$y <- dat$TMaleMinDistance    # response


full_model <- lmer(
  y ~ stage + (1 + stage | MaleID),
  data = dat,
  REML = TRUE
)


null_model <- lmer(
  y ~ stage +
    #independent intercepts and slopes
    (1 | MaleID) +
    (1 | BroodID) +
    (0 + stage | MaleID),
  data = dat,
  REML = TRUE
)

result <- identifiability_test_mixed(
  full_model = full_model,
  null_model = null_model,
  data = dat,
  n_sim = 1000,
  statistic = "correlation",
  focal_group = "MaleID",
  slope_term = "stage",
  labels = c("Breeding stage (centered)", "Aggressivness"),
  seed = 1,
  exclude_singular = FALSE
)

result$p_value
result$confint
result$observed_statistic
row1 = result$figure
```




```{r}
#code + data from: Jolles, Jolle; Briggs, Helen; Araya-Ajoy, Yimen; Boogert, Neeltje (2019), “Data for: Personality, plasticity and predictability in sticklebacks: bold fish are less plastic and more predictable than shy fish”, Mendeley Data, V2, doi: 10.17632/d9xz2k5pyk.2
library(lme4)
library(ggplot2)

dat <- read.csv("Sticklebacks/boldnessdat.csv")
head(dat)
dbef <- dat[!dat$testwk %in% c(10),]
dbef.list =
  list(Y          = dbef$prop.out,
       BATCH      = dbef$batch - 1,
       ID         = dbef$id,
       WEEK       = as.numeric(dbef$testwk),
       TRIAL      = as.numeric(dbef$trial) - 3,
       N          = length(dbef$prop.out),
       N_id       = length(unique(dbef$id)))
dbef.list$Y.z       <- (dbef.list$Y - mean(dbef.list$Y)) / sd(dbef.list$Y)
dbef.df = as.data.frame(dbef.list)
dbef.df$WEEK.c = dbef.df$WEEK-mean(dbef.df$WEEK)
full_model <- lmer(Y.z ~ WEEK + (WEEK | ID), dbef.df, 
                            control = lmerControl(optimizer = "bobyqa"))

null_model <- lmer(
  Y.z ~ WEEK +
    #independent intercepts and slopes
    (1 | ID) +
    (0 + WEEK | ID),
  data = dbef.df,
  REML = TRUE
)

result <- identifiability_test_mixed(
  full_model = full_model,
  null_model = null_model,
  data = dbef.df,
  n_sim = 1000,
  statistic = "correlation",
  focal_group = "ID",
  slope_term = "WEEK",
  labels = c("Week","Standardized boldness"),
  seed = 1,
  exclude_singular = FALSE,
  tags = c("D", "E", "F")
)



result$p_value
row2 = result$figure
result$observed_statistic
result$confint
```

```{r}
#centered
full_model <- lmer(Y.z ~ WEEK.c + (WEEK.c | ID), dbef.df, 
                            control = lmerControl(optimizer = "bobyqa"))

null_model <- lmer(
  Y.z ~ WEEK.c +
    #independent intercepts and slopes
    (1 | ID) +
    (0 + WEEK.c | ID),
  data = dbef.df,
  REML = TRUE
)
result2 <- identifiability_test_mixed(
  full_model = full_model,
  null_model = null_model,
  data = dbef.df,
  n_sim = 1000,
  statistic = "correlation",
  focal_group = "ID",
  slope_term = "WEEK.c",
  labels = c("Week (centered)","Standardized boldness"),
  seed = 1,
  exclude_singular = FALSE,
  tags = c("D", "E", "F")
)
```


```{r}
#Mussel sperm
#analysis is reconstructed from author stated methods
data3 = read.csv("Mussel_Sperm/Hadlow_et_al_2023_Female_reproductive_fluids.csv")
head(data3)


# Construct identifiers and predictors
library(FactoMineR)
library(dplyr)
library(stringr)
library(lme4)

dat <- data3 %>%
  mutate(
    male_id = factor(
      str_remove(animal_id, "\\.[1-4]\\.[ES]$")
    ),
    treatment = factor(
      str_extract(animal_id, "[ES]$"),
      levels = c("S", "E"),
      labels = c("Seawater", "Egg water")
    ),
    block_id = factor(egg_pool),

    # Already measured as minutes since spawning
    ejaculate_age = motility_time
  )

#calculate PC1

pca_vars <- c(
  "motile_mean_alh",
  "motile_mean_bcf",
  "motile_mean_lin",
  "motile_mean_str",
  "motile_mean_vap",
  "motile_mean_vcl",
  "motile_mean_vsl",
  "motile_mean_wob"
)

pca_result <- PCA(
  dat[, pca_vars],
  scale.unit = TRUE,
  ncp = 8,
  graph = FALSE
)

dat$PC1 <- pca_result$ind$coord[, 1]
#percentage variance explained agrees with the paper. Exact loadings are different, isnce PCA scale is arbitrary
pca_result$eig

#standardize ejaculate age
dat$age_z <- as.numeric(scale(dat$motility_time))
#filter to just seawater
sea_dat = subset(dat, treatment == "Seawater")
sea_dat <- subset(dat, treatment == "Seawater")

full_model <- lmer(
  PC1 ~ age_z +
    (1 + age_z | male_id) +
    (1 | block_id),
  data = sea_dat,
  REML = TRUE
)

null_model <- lmer(
  PC1 ~ age_z +
    #independent intercepts and slopes
    (1 | male_id) +
    (1 | block_id) +
    (0 + age_z | male_id),
  data = sea_dat,
  REML = TRUE
)

#run the test
result <- identifiability_test_mixed(
  full_model = full_model,
  null_model = null_model,
  data = sea_dat,
  n_sim = 1000,
  statistic = "correlation",
  focal_group = "male_id",
  slope_term = "age_z",
  labels = c("Standardized age","PC1 (motility)"),
  seed = 2,
  exclude_singular = FALSE,
  tags = c("G", "H", "I")
)
result$observed_statistic
result$p_value
result$confint
row3 = result$figure
```

```{r}
final_plot=row1/row2/row3
final_plot
```

~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
Power analysis (Figure S4)
```{r}
rm(list=ls())
source("identifiability_functions.R")
tradeoff_power_analysis(
GROUP_SAMPLE_SIZES = c(10,20,50,100,200,1000),
GENERATING_CORRELATIONS = c(-0.8, -0.5, -0.3, 0, 0.3, 0.5, 0.8),
# Means of the true intercept and slope distributions.
MEAN_INTERCEPT = 1.6,
MEAN_SLOPE     = -0.21,
#for the Soto data example, these are a conservative estimates, since it includes sampling/estimation variability
INTERCEPT_VARIANCE = 0.006,
SLOPE_VARIANCE     = 0.0018,
#Impossible to know what this truly is, but we can have an idea for order of magnitude
RESIDUAL_VARIANCE = 0.5,
#Centered soto data for example
PREDICTOR_VALUES = c(-1.5,-0.5,0.5,1.5),
REPLICATES_PER_PREDICTOR = 10,
N_OUTER_SIMULATIONS = 200,
N_INNER_NULL_SIMULATIONS = 200,
TEST_STATISTIC = "correlation",
SIGNIFICANCE_LEVEL = 0.05,
TEST_ALTERNATIVE = "two.sided",
CENTER_PREDICTOR = FALSE,
PREDICTOR_CENTER = NULL,
TEST_UNCENTERED_INTERCEPT = FALSE,
MASTER_SEED = 1,
RESULTS_FILE = "reaction_norm_power_results.rds",
SUMMARY_FILE = "reaction_norm_power_summary.csv",
FIGURE_FILE  = "reaction_norm_power_panel_A.png",
RESUME_EXISTING_RUN = FALSE,
VERBOSE = TRUE)
```
















